---
title: "plots, lots of plots"
author: "Melanie Smith"
date: "12 July 2023"
output: html_document
---

Now that we have a number of different models run within the preeclampsia SCOPE + STOP RNA-seq data (with and without covariates), this script is used to compare the model outputs such as logFC, p-value, sigma and t-statistic. Only model outputs related to the `featureCounts` v2.0.3 sex informed counts is considered.  

# Load required libraries
```{r load Libraries}
library(plyr)
library(reshape2)
library(dplyr)
library(tidyr)
library(stringr)
library(edgeR)
library(readxl)
library(readr)
library(magrittr)
library(ggplot2)
library(tibble)
# library(ggbeeswarm)
# library(ggrepel)
# library(RColorBrewer)
# library(pheatmap)
# library(ggpubr)

# set project directory
projectDir <- "/home/smit1924/sexInformedPreeclampsia"
# projectDir <- "/Users/pillmaka/Desktop/Melanie/drive-download"

# define input files
melanie_v2_0_3_filtered_inFile <- file.path(projectDir, "melanie_v2_0_3_output/melanie_v2_0_3filtered_DE.csv")
melanie_v2_0_3_pValue_inFile <- file.path(projectDir, "melanie_v2_0_3_output/melanie_v2_0_3pValue_DE.csv")
maternalAgefiltered_infile <- file.path(projectDir, "maternalAge_output/maternalAgefiltered_DE.csv")
maternalBMIfiltered_infile <- file.path(projectDir, "maternalBMI_output/maternalBMIfiltered_DE.csv")
maternalAgePvalue_infile <- file.path(projectDir, "maternalAge_output/maternalAgepValue_DE.csv")
maternalBMIPvalue_infile <- file.path(projectDir, "maternalBMI_output/maternalBMIpValue_DE.csv")
noCovariateDE_infile <- file.path(projectDir, "noCovariate_output/noCovariatefiltered_DE.csv")
noCovariatePvalue_infile <- file.path(projectDir, "noCovariate_output/noCovariatepValue_DE.csv")

# define out Dir
outPlotsDir <- file.path(projectDir, "clearBoxPlots/")

dir.create(outPlotsDir)
```

# Import DE files

```{r}
thesis_model_DE <- read_csv(file = melanie_v2_0_3_filtered_inFile)
thesis_model_pValue <- read_csv(file = melanie_v2_0_3_pValue_inFile)
maternalAge_model_DE <- read_csv(file = maternalAgefiltered_infile)
maternalAge_pValue <- read_csv(file = maternalAgePvalue_infile)
maternalBMI_model_DE <- read_csv(file = maternalBMIfiltered_infile)
maternalBMI_pValue <- read_csv(file = maternalBMIPvalue_infile)
noCovariate_model_DE <- read_csv(file = noCovariateDE_infile)
noCovariate_pValue <- read_csv(file = noCovariatePvalue_infile)
```

# Combine the tables (logFC, adj.P.Val and sigma retained)

```{r}
# DE using filtered counts (2CPM in 5 samples)
combined_LogFC <- dplyr::full_join(
  maternalAge_model_DE %>% dplyr::select(ensembl_gene_id, hgnc_symbol, sigma, logFC_F, logFC_M, adj.P.Val_F, adj.P.Val_M),
  maternalBMI_model_DE %>% dplyr::select(ensembl_gene_id, hgnc_symbol, sigma, logFC_F, logFC_M, adj.P.Val_F, adj.P.Val_M),
                                  by=c('hgnc_symbol', 'ensembl_gene_id'),
                                  suffix=c("_age", "_bmi")) %>%
  dplyr::full_join(., thesis_model_DE %>% dplyr::select(ensembl_gene_id, hgnc_symbol, sigma,
                                                        logFC_F_thesis = logFC_F,
                                                        logFC_M_thesis = logFC_M,
                                                        adj.P.Val_F_thesis = adj.P.Val_F,
                                                        adj.P.Val_M_thesis = adj.P.Val_M),
                   by=c('hgnc_symbol', 'ensembl_gene_id')) %>%
  dplyr::full_join(., noCovariate_model_DE %>% dplyr::select(ensembl_gene_id, hgnc_symbol, sigma,
                                                        logFC_F_noCovariate = logFC_F,
                                                        logFC_M_noCovariate = logFC_M,
                                                        adj.P.Val_F_noCovariate = adj.P.Val_F,
                                                        adj.P.Val_M_noCovariate = adj.P.Val_M),
                   by=c('hgnc_symbol', 'ensembl_gene_id'))

combined_pValue <- dplyr::full_join(maternalAge_pValue, maternalBMI_pValue,
                                    by=c('hgnc_symbol', 'ensembl_gene_id'),
                                    suffix=c("_age", "_bmi")) %>%
  dplyr::full_join(., thesis_model_pValue %>% dplyr::select(ensembl_gene_id, hgnc_symbol,
                                                        t_F_thesis = t_F,
                                                        P.Value_F_thesis = P.Value_F,
                                                        adj.P.Val_F_thesis = adj.P.Val_F,
                                                        t_M_thesis = t_M,
                                                        P.Value_M_thesis = P.Value_M,
                                                        adj.P.Val_M_thesis = adj.P.Val_M),
                   by=c('hgnc_symbol', 'ensembl_gene_id')) %>%
  dplyr::full_join(., noCovariate_pValue %>% dplyr::select(ensembl_gene_id, hgnc_symbol,
                                                        t_F_noCovariate = t_F,
                                                        P.Value_F_noCovariate = P.Value_F,
                                                        adj.P.Val_F_noCovariate = adj.P.Val_F,
                                                        t_M_noCovariate = t_M,
                                                        P.Value_M_noCovariate = P.Value_M,
                                                        adj.P.Val_M_noCovariate = adj.P.Val_M),
                   by=c('hgnc_symbol', 'ensembl_gene_id'))

```

# Scatter plot of p-Values

```{r}
# thesis model versus maternal age only
ggplot(combined_pValue,
       aes(x = -log10(P.Value_F_thesis),
           y = -log10(P.Value_F_age))) +
  geom_point() +
  xlim(0, max(-log10(combined_pValue$P.Value_F_thesis), -log10(combined_pValue$P.Value_F_age))) +
  ylim(0, max(-log10(combined_pValue$P.Value_F_thesis), -log10(combined_pValue$P.Value_F_age))) +
  geom_abline(intercept = 0, slope = 1, size = 0.5, colour="red") +
  ggtitle("Female: thesis model versus age only model (-log10(p.val))")
ggsave(filename = file.path(outPlotsDir, "neglog10PVal_thesis_age_F.png"),
       width = 7,
       height = 7,
       units = "in",
       dpi = 150)
# thesis model versus maternal bmi only
ggplot(combined_pValue,
       aes(x = -log10(P.Value_F_thesis),
           y = -log10(P.Value_F_bmi))) +
  geom_point() +
  xlim(0, max(-log10(combined_pValue$P.Value_F_thesis), -log10(combined_pValue$P.Value_F_bmi))) +
  ylim(0, max(-log10(combined_pValue$P.Value_F_thesis), -log10(combined_pValue$P.Value_F_bmi))) +
  geom_abline(intercept = 0, slope = 1, size = 0.5, colour="red") +
  ggtitle("Female: thesis model versus bmi only model (-log10(p.val))")
ggsave(filename = file.path(outPlotsDir, "neglog10PVal_thesis_bmi_F.png"),
       width = 7,
       height = 7,
       units = "in",
       dpi = 150)
# no covariate model versus maternal age only
ggplot(combined_pValue,
       aes(x = -log10(P.Value_F_noCovariate),
           y = -log10(P.Value_F_age))) +
  geom_point() +
  xlim(0, max(-log10(combined_pValue$P.Value_F_noCovariate), -log10(combined_pValue$P.Value_F_age))) +
  ylim(0, max(-log10(combined_pValue$P.Value_F_noCovariate), -log10(combined_pValue$P.Value_F_age))) +
  geom_abline(intercept = 0, slope = 1, size = 0.5, colour="red") +
  ggtitle("Female: no covariate model versus age only model (-log10(p.val))")
ggsave(filename = file.path(outPlotsDir, "neglog10PVal_noCovariate_age_F.png"),
       width = 7,
       height = 7,
       units = "in",
       dpi = 150)
# no covariate model versus maternal bmi only
ggplot(combined_pValue,
       aes(x = -log10(P.Value_F_noCovariate),
           y = -log10(P.Value_F_bmi))) +
  geom_point() +
  xlim(0, max(-log10(combined_pValue$P.Value_F_noCovariate), -log10(combined_pValue$P.Value_F_bmi))) +
  ylim(0, max(-log10(combined_pValue$P.Value_F_noCovariate), -log10(combined_pValue$P.Value_F_bmi))) +
  geom_abline(intercept = 0, slope = 1, size = 0.5, colour="red") +
  ggtitle("Female: no covariate model versus bmi only model (-log10(p.val))")
ggsave(filename = file.path(outPlotsDir, "neglog10PVal_noCovariate_bmi_F.png"),
       width = 7,
       height = 7,
       units = "in",
       dpi = 150)
```

# Histogram plot P-Value
## model includes maternal age and maternal BMI
```{r}
# plot p-value thesis model in females
ggplot(data = combined_pValue,
       aes(x = P.Value_F_thesis)) +
  geom_histogram(binwidth = 0.01) +
      theme_bw() +
  theme(text = element_text(size=18)) +
  theme(axis.text.x = element_text(colour = "black",
                                   face = "bold"),
        axis.text.y = element_text(colour = "black",
                                   face = "bold")) +
  theme(axis.title.x = element_text(colour = "black",
                                   face = "bold"),
        axis.title.y = element_text(colour = "black",
                                   face = "bold")) +
  ggtitle("p-value maternal age and BMI as covariate in females")
# plot p-value thesis model in males
ggplot(data = combined_pValue,
       aes(x = P.Value_M_thesis)) +
  geom_histogram(binwidth = 0.01) +
            theme_bw() +
  theme(text = element_text(size=18)) +
  theme(axis.text.x = element_text(colour = "black",
                                   face = "bold"),
        axis.text.y = element_text(colour = "black",
                                   face = "bold")) +
  theme(axis.title.x = element_text(colour = "black",
                                   face = "bold"),
        axis.title.y = element_text(colour = "black",
                                   face = "bold")) +
  ggtitle("p-value thesis model in males")
```

# Histogram plot P-Value
## model includes maternal age only
```{r}
# plot p-value maternal age as covariate in females
ggplot(data = combined_pValue,
       aes(x = P.Value_F_age)) +
  geom_histogram(binwidth = 0.01) +
    theme_bw() +
  theme(text = element_text(size=18)) +
  theme(axis.text.x = element_text(colour = "black",
                                   face = "bold"),
        axis.text.y = element_text(colour = "black",
                                   face = "bold")) +
  theme(axis.title.x = element_text(colour = "black",
                                   face = "bold"),
        axis.title.y = element_text(colour = "black",
                                   face = "bold")) +
  ggtitle("p-value maternal age as covariate in females")
# plot p-value maternal age as covariate in males
ggplot(data = combined_pValue,
       aes(x = P.Value_M_age)) +
  geom_histogram(binwidth = 0.01) +
          theme_bw() +
  theme(text = element_text(size=18)) +
  theme(axis.text.x = element_text(colour = "black",
                                   face = "bold"),
        axis.text.y = element_text(colour = "black",
                                   face = "bold")) +
  theme(axis.title.x = element_text(colour = "black",
                                   face = "bold"),
        axis.title.y = element_text(colour = "black",
                                   face = "bold")) +
  ggtitle("p-value maternal age as covariate in males")
```

## model includes maternal BMI only
```{r}
# plot p-value maternal bmi as covariate in females
ggplot(data = combined_pValue,
       aes(x = P.Value_F_bmi)) +
  geom_histogram(binwidth = 0.01) +
        theme_bw() +
  theme(text = element_text(size=18)) +
  theme(axis.text.x = element_text(colour = "black",
                                   face = "bold"),
        axis.text.y = element_text(colour = "black",
                                   face = "bold")) +
  theme(axis.title.x = element_text(colour = "black",
                                   face = "bold"),
        axis.title.y = element_text(colour = "black",
                                   face = "bold")) +
  ggtitle("p-value maternal BMI as covariate in females")

# plot p-value maternal bmi as covariate in males
ggplot(data = combined_pValue,
       aes(x = P.Value_M_bmi)) +
  geom_histogram(binwidth = 0.01) +
              theme_bw() +
  theme(text = element_text(size=18)) +
  theme(axis.text.x = element_text(colour = "black",
                                   face = "bold"),
        axis.text.y = element_text(colour = "black",
                                   face = "bold")) +
  theme(axis.title.x = element_text(colour = "black",
                                   face = "bold"),
        axis.title.y = element_text(colour = "black",
                                   face = "bold")) +
  ggtitle("p-value maternal bmi as covariate in males")
```

## model includes no covariates
```{r}
# plot p-value maternal bmi as covariate in females
ggplot(data = combined_pValue,
       aes(x = P.Value_F_noCovariate)) +
          theme_bw() +
  theme(text = element_text(size=18)) +
  theme(axis.text.x = element_text(colour = "black",
                                   face = "bold"),
        axis.text.y = element_text(colour = "black",
                                   face = "bold")) +
  theme(axis.title.x = element_text(colour = "black",
                                   face = "bold"),
        axis.title.y = element_text(colour = "black",
                                   face = "bold")) +
  geom_histogram(binwidth = 0.01) +
  
  ggtitle("p-value no covariate in females")

# plot p-value maternal bmi as covariate in males
ggplot(data = combined_pValue,
       aes(x = P.Value_M_noCovariate)) +
  geom_histogram(binwidth = 0.01) +
            theme_bw() +
  theme(text = element_text(size=18)) +
  theme(axis.text.x = element_text(colour = "black",
                                   face = "bold"),
        axis.text.y = element_text(colour = "black",
                                   face = "bold")) +
  theme(axis.title.x = element_text(colour = "black",
                                   face = "bold"),
        axis.title.y = element_text(colour = "black",
                                   face = "bold")) +
  ggtitle("p-value no covariate in males")
```

# Histogram plot t-statistic
## model includes maternal age only
```{r}
# plot t-statistic maternal age as covariate in females
ggplot(data = combined_pValue,
       aes(x = t_F_age)) +
  geom_histogram(binwidth = 0.1) +
  ggtitle("t-statistic maternal age as covariate in females")
# plot t-statistic maternal age as covariate in males
ggplot(data = combined_pValue,
       aes(x = t_M_age)) +
  geom_histogram(binwidth = 0.1) +
  ggtitle("t-statistic maternal age as covariate in males")
```

## model includes maternal bmi only
```{r}
# plot t-statistic maternal age as covariate in females
ggplot(data = combined_pValue,
       aes(x = t_F_bmi)) +
  geom_histogram(binwidth = 0.1) +
  ggtitle("t-statistic maternal BMI as covariate in females")
# plot t-statistic maternal bmi as covariate in males
ggplot(data = combined_pValue,
       aes(x = t_M_bmi)) +
  geom_histogram(binwidth = 0.1) +
  ggtitle("t-statistic maternal BMI as covariate in males")
```

# plot adjusted p-value
## no covariate model compared to age only model
```{r}
# plot logged adjusted.p.value no covariate versus age only in females
ggplot(combined_LogFC,
       aes(x = -log10(adj.P.Val_F_noCovariate),
           y = -log10(adj.P.Val_F_age))) +
  geom_point() +
  ggtitle("Female: No covariate versus age only model (-log10(adj.p.val))")
ggsave(filename = file.path(outPlotsDir, "neglog10AdjPVal_noCovariate_age_F.png"),
       width = 7,
       height = 7,
       units = "in",
       dpi = 150)
# plot logged adjusted.p.value no covariate versus age only in males
ggplot(combined_LogFC,
       aes(x = -log10(adj.P.Val_M_noCovariate),
           y = -log10(adj.P.Val_M_age))) +
  geom_hline(yintercept=1.30103, linetype="dashed", color = "red") +
  geom_vline(xintercept=1.30103, linetype="dashed", color = "red") +
  geom_point() +
  xlim(2.5, 0) +
  ylim(2.5, 0) +
  geom_abline(intercept = 0, slope = 1, size = 0.5, colour="red") +
  ggtitle("Male: No covariate versus age only model (-log10(adj.p.val))")
ggsave(filename = file.path(outPlotsDir, "neglog10AdjPVal_noCovariate_age_M.png"),
       width = 7,
       height = 7,
       units = "in",
       dpi = 150)
```

## no covariate model compared to bmi only model
```{r}
# plot logged adjusted.p.value no covariate versus bmi only in females
ggplot(combined_LogFC,
       aes(x = -log10(adj.P.Val_F_noCovariate),
           y = -log10(adj.P.Val_F_bmi))) +
  geom_point() +
  ggtitle("Female: No covariate versus bmi only model (-log10(adj.p.val))")
ggsave(filename = file.path(outPlotsDir, "neglog10AdjPVal_noCovariate_bmi_F.png"),
       width = 7,
       height = 7,
       units = "in",
       dpi = 150)
# plot logged adjusted.p.value no covariate versus bmi only in males
ggplot(combined_LogFC,
       aes(x = -log10(adj.P.Val_M_noCovariate),
           y = -log10(adj.P.Val_M_bmi))) +
  geom_hline(yintercept=1.30103, linetype="dashed", color = "red") +
  geom_vline(xintercept=1.30103, linetype="dashed", color = "red") +
  geom_point() +
  xlim(2.5, 0) +
  ylim(2.5, 0) +
  geom_abline(intercept = 0, slope = 1, size = 0.5, colour="red") +
  ggtitle("Male: No covariate versus bmi only model (-log10(adj.p.val))")
ggsave(filename = file.path(outPlotsDir, "neglog10AdjPVal_noCovariate_bmi_M.png"),
       width = 7,
       height = 7,
       units = "in",
       dpi = 150)

```

## model with onyl maternalAge compared to thesis model
```{r}
# plot logged adjusted.p.value maternal age versus thesis in females
ggplot(combined_LogFC,
       aes(x = -log10(adj.P.Val_F_thesis),
           y = -log10(adj.P.Val_F_age))) +
  geom_point() +
  ggtitle("Female: Age only versus thesis model (-log10(adj.p.val))")
ggsave(filename = file.path(outPlotsDir, "neglog10AdjPVal_age_thesis_F.png"),
       width = 7,
       height = 7,
       units = "in",
       dpi = 150)
# plot logged adjusted.p.value maternal age versus thesis in males
ggplot(combined_LogFC,
       aes(x = -log10(adj.P.Val_M_thesis),
           y = -log10(adj.P.Val_M_age))) +
  geom_hline(yintercept=1.30103, linetype="dashed", color = "red") +
  geom_vline(xintercept=1.30103, linetype="dashed", color = "red") +
  geom_point() +
  xlim(2.5, 0) +
  ylim(2.5, 0) +
  geom_abline(intercept = 0, slope = 1, size = 0.5, colour="red") +
  ggtitle("Male: Age only versus thesis model (-log10(adj.p.val))")
ggsave(filename = file.path(outPlotsDir, "neglog10AdjPVal_age_thesis_M.png"),
       width = 7,
       height = 7,
       units = "in",
       dpi = 150)

# png(file=file.path(outPlotsDir, "heatmap_sig_AdjPVal_M.png"),
#     width=12,
#     height=8,
#     units = "in",
#     res = 150)
# pheatmap(dplyr::filter(combined_LogFC, adj.P.Val_M_bmi < 0.05 | adj.P.Val_M_thesis < 0.05 & 0.05 | adj.P.Val_M_age < 0.05) %>%
#            dplyr::select(hgnc_symbol, adj.P.Val_M_thesis, adj.P.Val_M_age, adj.P.Val_M_bmi) %>%
#            tibble::column_to_rownames("hgnc_symbol"),
#          cluster_rows = T,
#          cluster_cols = F,
#          show_rownames = TRUE, 
#          color = c("green", "grey"),
#          breaks = c(0, 0.05, .11),  # distances 0 to 0.05 are green, 0.05 to 1.1 grey
#          main = 'Male sig p-value')
# dev.off()

```

## model with onyl maternal BMI compared to thesis model
```{r}
# plot logged adjusted.p.value maternal bmi versus thesis in females
ggplot(combined_LogFC,
       aes(x = -log10(adj.P.Val_F_thesis),
           y = -log10(adj.P.Val_F_bmi))) +
  geom_point() +
  ggtitle("Female: BMI only versus thesis model (-log10(adj.p.val))")
ggsave(filename = file.path(outPlotsDir, "neglog10AdjPVal_bmi_thesis_F.png"),
       width = 7,
       height = 7,
       units = "in",
       dpi = 150)
# plot logged adjusted.p.value maternal bmi versus thesis in males
ggplot(combined_LogFC,
       aes(x = -log10(adj.P.Val_M_thesis),
           y = -log10(adj.P.Val_M_bmi))) +
    geom_hline(yintercept=1.30103, linetype="dashed", color = "red") +
  geom_vline(xintercept=1.30103, linetype="dashed", color = "red") +
  geom_point() +
  xlim(2.5, 0) +
  ylim(2.5, 0) +
  geom_abline(intercept = 0, slope = 1, size = 0.5, colour="red") +
  ggtitle("Male: BMI only versus thesis model (-log10(adj.p.val))")
ggsave(filename = file.path(outPlotsDir, "neglog10AdjPVal_bmi_thesis_M.png"),
       width = 7,
       height = 7,
       units = "in",
       dpi = 150)


```

# plot logFC
## Model with only maternaAge compared to thesis model
```{r}
# male age only versus thesis logFC
ggplot(data = combined_LogFC,
       aes(x = logFC_M_age,
           y = logFC_M_thesis)) +
  geom_point() +
  xlim(-7, 7) +
  ylim(-7, 7) +
  ggtitle("Male: age only versus thesis model (log2FC)")
ggsave(filename = file.path(outPlotsDir, "logFC_age_thesis_M.png"),
       width = 7,
       height = 7,
       units = "in",
       dpi = 150)
# female age only versus thesis logFC
ggplot(data = combined_LogFC,
       aes(x = logFC_F_age,
           y = logFC_F_thesis)) +
  geom_point() +
  xlim(-7, 7) +
  ylim(-7, 7) +
  ggtitle("Female: age only versus thesis model (log2FC)")
ggsave(filename = file.path(outPlotsDir, "logFC_age_thesis_F.png"),
       width = 7,
       height = 7,
       units = "in",
       dpi = 150)

```

## Model with only maternaBMI compared to thesis model
```{r}
# male BMI only versus thesis logFC
ggplot(data = combined_LogFC,
       aes(x = logFC_M_bmi,
           y = logFC_M_thesis)) +
  geom_point() +
  xlim(-7, 7) +
  ylim(-7, 7) +
  ggtitle("Male: bmi only versus thesis model (log2FC)")
ggsave(filename = file.path(outPlotsDir, "logFC_bmi_thesis_M.png"),
       width = 7,
       height = 7,
       units = "in",
       dpi = 150)
# female BMI only versus thesis logFC
ggplot(data = combined_LogFC,
       aes(x = logFC_F_bmi,
           y = logFC_F_thesis)) +
  geom_point() +
  xlim(-7, 7) +
  ylim(-7, 7) +
  ggtitle("Female: bmi only versus thesis model (log2FC)")
ggsave(filename = file.path(outPlotsDir, "logFC_bmi_thesis_F.png"),
       width = 7,
       height = 7,
       units = "in",
       dpi = 150)
```

## Plot model sigma
```{r}
# male BMI only versus thesis logFC
ggplot(data = combined_LogFC,
       aes(x = sigma,
           y = sigma_age)) +
  geom_point() +
  xlim(0, 10) +
  ylim(0, 10) +
  ggtitle("sigma: thesis v age only model")
ggsave(filename = file.path(outPlotsDir, "sigma_thesis_age.png"),
       width = 7,
       height = 7,
       units = "in",
       dpi = 150)
# female BMI only versus thesis logFC
ggplot(data = combined_LogFC,
       aes(x = sigma,
           y = sigma_bmi)) +
  geom_point() +
  xlim(0, 10) +
  ylim(0, 10) +
  ggtitle("sigma: thesis v bmi only model")
ggsave(filename = file.path(outPlotsDir, "sigma_thesis_bmi.png"),
       width = 7,
       height = 7,
       units = "in",
       dpi = 150)
```

# plot logFC for top 200 against covariate coefficiant
```{r}
# plot logFC against age coefficient
ggplot(data = dplyr::arrange(thesis_model_DE, adj.P.Val_M)[1:200,],
       aes(x = logFC_M,
           y = maternalAge)) +
  geom_point() +
  geom_smooth(method=lm) +
    ggtitle("Maternal Age Coefficient as a function of logFC (top 200 genes)")
# plot logFC against bmi coefficient
ggplot(data = dplyr::arrange(thesis_model_DE, adj.P.Val_M)[1:200,],
       aes(x = logFC_M,
           y = BMI)) +
  geom_point() +
  geom_smooth(method=lm) +
    ggtitle("Maternal BMI Coefficient as a function of logFC (top 200 genes)")

```

# Session information

```{r session info}

sessionInfo()

```
