---
title: "clearBoxDE"
author: "Melanie Smith"
date: "11 July 2023"
output: html_document
---

# Load required libraries
```{r load Libraries}
library(plyr)
library(reshape2)
library(dplyr)
library(stringr)
library(edgeR)
library(readxl)
library(readr)
library(magrittr)
library(tibble)
library(ggplot2)
library(ggrepel)
library(pheatmap)

# set project directory
projectDir <- "/home/smit1924/sexInformedPreeclampsia"
# projectDir <- "/Users/pillmaka/Desktop/Melanie/drive-download"

# expt_name <- 'maternalAge'
# expt_name <- 'maternalBMI'
expt_name <- 'noCovariate'
# expt_name <- 'interaction'

counts_geneID_file <- file.path(projectDir, "clearBoxRawData/rawCounts_melanie_v2.0.3.csv")

gene_info_file <- file.path(projectDir, "clearBoxRawData/gene_id_vs_symbol.csv")
metadata_file <- file.path(projectDir, "clearBoxRawData/cleanSampleMetadata.csv")

subsetGeneList_file <- file.path(projectDir, "clearBoxRawData/subsetGeneList.csv")
outDesignMatrix_file <- file.path(projectDir, paste0("clearBoxRawData/", expt_name, "_design.csv"))

outdir <- file.path(projectDir, paste0(expt_name, "_output"))
output_spreadsheet_file <- file.path(outdir, paste0(expt_name, "filtered_DE.csv"))
output_pValue_file <- file.path(outdir, paste0(expt_name, "pValue_DE.csv"))
output_log2CPM_file <- file.path(outdir, paste0(expt_name, "log2CPM.csv"))

dir.create(outdir)

# set max digits
options(digits=3)

# set filtering criteria
filterCPM <- 2
numSamples <- 5

```

# Import all input data tables
```{r}
## Count data
# import the count file for DE
rawCounts <- read_csv(file = counts_geneID_file)

## Metadata
metadata <- read_csv(file = metadata_file)

## Gene ID versus GeneSymbol
gene_info <- read_csv(file = gene_info_file)

## Gene ID versus GeneSymbol
gene_info <- read_csv(file = gene_info_file)
gene_info <- gene_info[!duplicated(gene_info$ensembl_gene_id), ]
```

# Establish DGElist object
```{r}
# drop the samples not required for this analysis
counts <- rawCounts %>%
  tibble::column_to_rownames("Geneid")
counts <- counts[, metadata$samplename]
# drop rows with zero in all columns
counts <- counts[rowSums(counts[])>0,] %>%
  tibble::rownames_to_column("ensembl_gene_id")
# put the ensemble gene IDs into column 1
counts <- dplyr::left_join(counts, gene_info, by = "ensembl_gene_id") %>%
  dplyr::select(., ensembl_gene_id, hgnc_symbol, everything())
dim(counts)
# establish the DGEList
dge_list_obj <- DGEList(counts = counts %>% tibble::column_to_rownames("ensembl_gene_id") %>% dplyr::select(., -hgnc_symbol),
                           samples = metadata,
                           group = metadata$group,
                           genes = counts %>% dplyr::select(ensembl_gene_id, hgnc_symbol))

```

# Filter biological noise and plot
```{r}

# new df of unfiltered cpm for the reduced DGEList
preFilterCPM <- cpm(dge_list_obj)

# remove low expressed genes (filterCPM in numSamples PE)
keep.exprs <- rowSums(preFilterCPM > filterCPM) >= numSamples

# Perform the filtering step and recalculate the TMM normalisation factors for each library.
dge_list_obj <- dge_list_obj[keep.exprs,,keep.lib.sizes = FALSE]

dim(dge_list_obj)

```

# TMM normalisation
```{r}
# calculate normalisation factors and apply to the DGEList object
dge_list_obj <- calcNormFactors(dge_list_obj,
                           method = "TMM")

# # set the new factor levels
sex_outcome_levels <- c("F_Control", "M_Control", "F_PE", "M_PE")
dge_list_obj$samples$sex_outcome <- factor(dge_list_obj$samples$sex_outcome, levels=sex_outcome_levels)
```

# Calculate log2CPM and save table as .csv

```{r}
log2CPM <- edgeR::cpm(dge_list_obj, log = TRUE)

# edgeR::cpm(dge_list_obj, log = TRUE) %>%
#   as.data.frame() %>%
#   tibble::rownames_to_column("ensembl") %>%
#   write_csv(., file = output_log2CPM_file,
#            col_names = TRUE)

```

# Differential Expression
- covariates are either  
  - maternal age  
  - maternal BMI  
```{r}
## No covariates
#Create and write out design matrix
model_noCovariate <- '~0 + sex_outcome'
# establish the design matrix
design_sex_PE <- model.matrix(formula(model_noCovariate),
                              data = dge_list_obj$samples)
# make the column names a little nicer
colnames(design_sex_PE) <- c(sex_outcome_levels)


## maternalAge only as covariate
#Create and write out design matrix
# model_maternalAge <- '~0 + sex_outcome + maternalAge'
# establish the design matrix
# design_sex_PE <- model.matrix(formula(model_maternalAge),
#                               data = dge_list_obj$samples)
# make the column names a little nicer
# colnames(design_sex_PE) <- c(sex_outcome_levels, "maternalAge")

## BMI only as covariate
#Create and write out design matrix
# model_maternalBMI <- '~0 + sex_outcome + BMI'
# establish the design matrix
# design_sex_PE <- model.matrix(formula(model_maternalBMI),
#                               data = dge_list_obj$samples)
# make the column names a little nicer
# colnames(design_sex_PE) <- c(sex_outcome_levels, "BMI")


# save a csv of the design matrix
# write_csv(data.frame(design_sex_PE) %>%
#             tibble::rownames_to_column("samplename"),
#           file = outDesignMatrix_file,
#           col_names = TRUE)
```

#Read in design matrix 
```{r}
design_matrix <- read_csv(outDesignMatrix_file)

#View what dtypes are there
design_matrix %>% as_tibble() %>% head()

design_matrix <- design_matrix %>% 
  column_to_rownames(var="samplename")

design_matrix
```

```{r}
# perform the voom
voom_sex_PE <- voom(dge_list_obj, design_matrix, plot = TRUE)

# fit the linear model
fit_sex_PE <- lmFit(voom_sex_PE, design_matrix)
```

```{r}
# set the contrasts - this makes it easier to "see" what we're testing
contrast_sex_PE <- makeContrasts(female_PE = F_PE-F_Control,
                                 male_PE = M_PE-M_Control,
                                 levels = design_matrix)
contrast_sex_PE

# fit a linear regression to the contrast questions
contrast_fit_sex_PE <- contrasts.fit(fit_sex_PE, contrast_sex_PE)

# perform bayesian adjustment
contrast_fit_sex_PE <- eBayes(contrast_fit_sex_PE)
```

```{r}
# summary table of the sex PE fit
summary(decideTests(contrast_fit_sex_PE,
                    adjust.method = "fdr",
                    p.value = 0.05))

# all DE results for the female_PE comparison
allTable_female_PE <- topTable(contrast_fit_sex_PE,
                             coef = 1,
                             n = Inf,
                             sort = "p")

# all DE results for the male comparison
allTable_male_PE <- topTable(contrast_fit_sex_PE,
                             coef = 2,
                             n = Inf,
                             sort = "p")

# write out a few summary tables
# df <- left_join(allTable_female_PE %>% dplyr::select(-t, -P.Value, -B),
#                allTable_male_PE %>% dplyr::select(-t, -P.Value, -B),
#                by=c('hgnc_symbol', 'ensembl_gene_id'),
#           suffix=c("_F", "_M"))
# 
# coeff_and_sigma_df <- cbind(fit_sex_PE$coefficients, sigma=fit_sex_PE$sigma) %>%
#   as.data.frame() %>%
#   rownames_to_column(var="ensembl_gene_id")
# 
# df <- left_join(coeff_and_sigma_df,
#                df,
#                by="ensembl_gene_id")
# 
# write_csv(df %>% relocate(hgnc_symbol, .after=ensembl_gene_id),
#           file=output_spreadsheet_file)
# 
# df2 <- left_join(allTable_female_PE %>% dplyr::select(ensembl_gene_id, hgnc_symbol, t, P.Value, adj.P.Val),
#                allTable_male_PE %>% dplyr::select(ensembl_gene_id, hgnc_symbol, t, P.Value, adj.P.Val),
#                by=c('hgnc_symbol', 'ensembl_gene_id'),
#           suffix=c("_F", "_M"))
# 
# write_csv(df2 %>% relocate(hgnc_symbol, .after=ensembl_gene_id),
#           file=output_pValue_file)

```

## Volcano plot for visualisation of differential expression
### male

```{r volcano plots male}

decide <- decideTests(contrast_fit_sex_PE)
d2 <- topTable(contrast_fit_sex_PE, coef = 2, n = Inf, sort = "p") #[,c(2,6)]
d2$threshold <- 0
d2$threshold[0:nrow(filter(allTable_male_PE, adj.P.Val < 0.05))] <- 1
d2$threshold <- as.factor(d2$threshold)
d3 <- decideTests(contrast_fit_sex_PE) %>%
  as.data.frame()
d3 <- d3[2]
d4 <- left_join(tibble::rownames_to_column(d2), (tibble::rownames_to_column(d3)),
                       by = "rowname")
d4$male_PE <- as.factor(d4$male_PE)
d4$neg.log10FDR <- -log10(d4$adj.P.Val)
# set colours for the volcano plot
colour <- c("red", "black", "darkgreen")
FDR_FC_Decide <- decideTests(contrast_fit_sex_PE, lfc = 0)[,"male_PE"] %>%
  as.data.frame() %>%
  set_colnames("FDR_FC_Decide") %>%
  tibble::rownames_to_column() %>%
  left_join(., d4, by = "rowname")
FDR_FC_Decide$FDR_FC_Decide <- as.factor(FDR_FC_Decide$FDR_FC_Decide)

volcano_male <- ggplot(data = FDR_FC_Decide,
                  aes(x = logFC,
                      y = neg.log10FDR,
                      label = rowname,
                      colour = FDR_FC_Decide)) +
  geom_point(alpha=0.9, size=2.00) +
  xlab(expression(Log[2]*" Fold Change")) + ylab(expression("-ve "*Log[10]*" FDR")) +
  scale_color_manual(values = colour, name="mRNA\nRegulation",
                         breaks = c("-1", "0", "1"),
                         labels = c("Down-regulated", "Not Significant", "Up-regulated")) +
  theme(plot.title = element_text(hjust = 0.5)) +
  geom_hline(yintercept = -log10(.05), linetype = "dotted") +
  geom_vline(xintercept = c(-1, 1), linetype = "dotted") +
  theme_bw() +
  theme(text = element_text(size=18)) +
  theme(axis.text.x = element_text(colour = "black",
                                   face = "bold"),
        axis.text.y = element_text(colour = "black",
                                   face = "bold")) +
  theme(axis.title.x = element_text(colour = "black",
                                   face = "bold"),
        axis.title.y = element_text(colour = "black",
                                   face = "bold")) +
  geom_label_repel(data = dplyr::filter(FDR_FC_Decide, FDR_FC_Decide==1),
                  aes(label = hgnc_symbol),
                  fill = "white",
                  xlim = c(7, 0),
                  max.overlaps = Inf,
                  show.legend = FALSE) +
  geom_label_repel(data = dplyr::filter(FDR_FC_Decide, FDR_FC_Decide==-1),
                  aes(label = hgnc_symbol),
                  fill = "white",
                  xlim = c(-7, 0),
                  max.overlaps = Inf,
                  show.legend = FALSE) +
  theme(legend.position="bottom")

ggsave(plot = volcano_male,
       filename = file.path(projectDir, "plots/volcano_male.png"),
       units = "in",
       width = 12,
       height = 12,
       dpi = 150)



```

# Heatmaps
```{r}
# Modify ordering of the clusters using clustering callback option
callback = function(hc, mat){
    sv = svd(t(mat))$v[,1]
    dend = reorder(as.dendrogram(hc), wts = sv)
    as.hclust(dend)
}

annotation_cols <- metadata[, c("samplename", "Sex", "Outcome")] %>%
  tibble::column_to_rownames("samplename")

scaled_exp <- scale(t(subset(log2CPM, rownames(log2CPM) %in% dplyr::filter(FDR_FC_Decide, threshold == 1)$ensembl_gene_id)))

# png(file=file.path(outdir, "heatmap_scaled_sig_AdjPVal_M.png"),
#     width=12,
#     height=8,
#     units = "in",
#     res = 150)
pheatmap(t(scaled_exp),
         annotation_col = annotation_cols,
         show_rownames = FALSE,
         show_colnames = TRUE,
         cutree_cols = 3,
         clustering_callback = callback)
# dev.off()

```

# Session information
```{r session info}

sessionInfo()

```