---
title: "clearBoxDE"
author: "Melanie Smith"
date: "11 July 2023"
output: html_document
---

# Load required libraries
```{r load Libraries}
library(plyr)
library(reshape2)
library(dplyr)
library(stringr)
library(edgeR)
library(readxl)
library(readr)
library(magrittr)
# library(ggplot2)
library(tibble)
# library(ggbeeswarm)
# library(ggrepel)
# library(RColorBrewer)
# library(pheatmap)
# library(ggpubr)

# set project directory
projectDir <- "/home/smit1924/sexInformedPreeclampsia"
# projectDir <- "/Users/pillmaka/Desktop/Melanie/drive-download"

expt_name <- 'jimmy'
counts_geneID_file <- file.path(projectDir, "clearBoxRawData/jimmy_counts_geneID.csv")

# expt_name <- 'melanie_oldpipeline'
# counts_geneID_file <- file.path(projectDir, "clearBoxRawData/rawCounts_melanie_v1.6.4.csv")
 
# expt_name <- 'melanie_v2_0_3'
# counts_geneID_file <- file.path(projectDir, "clearBoxRawData/rawCounts_melanie_v2.0.3.csv")

gene_info_file <- file.path(projectDir, "clearBoxRawData/gene_id_vs_symbol.csv")
metadata_file <- file.path(projectDir, "clearBoxRawData/cleanSampleMetadata.csv")

subsetGeneList_file <- file.path(projectDir, "clearBoxRawData/subsetGeneList.csv")
outDesignMatrix_file <- file.path(projectDir, "clearBoxRawData/design.csv")

outdir <- file.path(projectDir, paste0(expt_name, "_output"))
output_spreadsheet_file <- file.path(outdir, paste0(expt_name, "_DE.csv"))

dir.create(outdir)

# set max digits
options(digits=3)

```

# Import all input data tables
```{r}
## Count data
# import the count file for DE
rawCounts <- read_csv(file = counts_geneID_file)

## Metadata
metadata <- read_csv(file = metadata_file)

## Gene ID versus GeneSymbol
gene_info <- read_csv(file = gene_info_file)

## Subset gene list
subsetGeneList <- read_csv(file = subsetGeneList_file)

```

# Prepare counts for DGElist
- Here we will take the count table and keep only genes with an ensembl ID in the thesis DE  
- Then, using the metadata table, ensure only the samples used in the final DE are present and in the same order as the metadata$samplename column  
- Then, replace the ensembl IDs with gene symbols and put the gene symbols into the rowname  

```{r}
dim(rawCounts)

# filter the counts to keep only genes in the final thesis DE
filteredCounts <- subset(rawCounts, Geneid %in% subsetGeneList$Geneid) %>%
  tibble::column_to_rownames("Geneid")
dim(filteredCounts)

# subset the counts to keep only samples in the final DE
filteredCounts <- dplyr::select(filteredCounts, all_of(metadata$samplename))
dim(filteredCounts)

```

# Establish DGElist object

```{r}

# put the ensemble gene IDs into column 1
counts <- filteredCounts %>%
  tibble::rownames_to_column("ensembl_gene_id") %>%
  left_join(., gene_info, by = "ensembl_gene_id") %>%
  dplyr::select(., ensembl_gene_id, hgnc_symbol, everything())

dim(counts)

# establish the DGEList
dge_list_obj <- DGEList(counts = counts %>% tibble::column_to_rownames("ensembl_gene_id") %>% dplyr::select(., -hgnc_symbol),
                           samples = metadata,
                           group = metadata$group,
                           genes = counts %>% dplyr::select(ensembl_gene_id, hgnc_symbol))

```

# TMM normalisation

```{r}
# calculate normalisation factors and apply to the DGEList object
dge_list_obj <- calcNormFactors(dge_list_obj,
                           method = "TMM")

# # set the new factor levels
sex_outcome_levels <- c("F_Control", "M_Control", "F_PE", "M_PE")
dge_list_obj$samples$sex_outcome <- factor(dge_list_obj$samples$sex_outcome, levels=sex_outcome_levels)
```

# Differential Expression
- covariates  
  - maternal age  
  - maternal BMI  
  
```{r}
# #Create and write out design matrix
# 
# # establish the design matrix
# design_sex_PE <- model.matrix(~0 + sex_outcome + maternalAge + BMI,
#                        data = dge_list_obj$samples)
# 
# # make the column names a little nicer
# colnames(design_sex_PE) <- c(sex_outcome_levels,  c("maternalAge", "BMI"))
# 
# # save a csv of the design matrix
# write_csv(data.frame(design_sex_PE) %>%
#             tibble::rownames_to_column("samplename"),
#           file = outDesignMatrix_file,
#           col_names = TRUE)
```

#Read in design matrix 
```{r}
design_matrix <- read_csv(outDesignMatrix_file)

#View what dtypes are there
design_matrix %>% as_tibble() %>% head()

design_matrix <- design_matrix %>% 
  column_to_rownames(var="samplename")

design_matrix
```


```{r}
# perform the voom
voom_sex_PE <- voom(dge_list_obj, design_matrix, plot = TRUE)

# fit the linear model
fit_sex_PE <- lmFit(voom_sex_PE, design_matrix)
```

```{r}
# set the contrasts - this makes it easier to "see" what we're testing
contrast_sex_PE <- makeContrasts(female_PE = F_PE-F_Control,
                                 male_PE = M_PE-M_Control,
                                 levels = design_matrix)
contrast_sex_PE

# fit a linear regression to the contrast questions
contrast_fit_sex_PE <- contrasts.fit(fit_sex_PE, contrast_sex_PE)

# perform bayesian adjustment
contrast_fit_sex_PE <- eBayes(contrast_fit_sex_PE)
```

```{r}
# summary table of the sex PE fit
summary(decideTests(contrast_fit_sex_PE,
                    adjust.method = "fdr",
                    p.value = 0.05))

# all DE results for the female_PE comparison
allTable_female_PE <- topTable(contrast_fit_sex_PE,
                             coef = 1,
                             n = Inf,
                             sort = "p")

# all DE results for the male comparison
allTable_male_PE <- topTable(contrast_fit_sex_PE,
                             coef = 2,
                             n = Inf,
                             sort = "p")

df <- left_join(allTable_female_PE %>% dplyr::select(-t, -P.Value, -B), 
               allTable_male_PE %>% dplyr::select(-t, -P.Value, -B), 
               by=c('hgnc_symbol', 'ensembl_gene_id'), 
          suffix=c("_F", "_M"))

coeff_and_sigma_df <- cbind(fit_sex_PE$coefficients, sigma=fit_sex_PE$sigma) %>% 
  as.data.frame() %>% 
  rownames_to_column(var="ensembl_gene_id")

df <- left_join(coeff_and_sigma_df, 
               df,
               by="ensembl_gene_id")

write_csv(df %>% relocate(hgnc_symbol, .after=ensembl_gene_id), 
          file=output_spreadsheet_file)

```

# Session information

```{r session info}

sessionInfo()

```