---
title: "clearBox_DE_cibersort_prop_cov"
author: "Melanie Smith"
date: "4 March 2024"
output: html_document
---

# Load required libraries

```{r load Libraries}
library(plyr)
library(reshape2)
library(dplyr)
library(stringr)
library(edgeR)
library(readxl)
library(readr)
library(magrittr)
library(tibble)
library(ggplot2)
library(ggrepel)
library(pheatmap)

# set project directory
projectDir <- "/home/smit1924/preeclampsia_sex_chromosome_informed"
# projectDir <- "/Users/pillmaka/Desktop/Melanie/drive-download"

expt_name <- 'cibersort_Covariate'


counts_geneID_file <- file.path(projectDir, "clearBoxRawData/rawCounts_melanie_v2.0.3.csv")
dge_list_obj_file <- file.path(projectDir, "noCovariate_output/DGEList_filtered_normalised.rds")
allTable_female_file <- file.path(projectDir, "noCovariate_output/allTable_female.csv")
allTable_male_file <- file.path(projectDir, "noCovariate_output/allTable_male.csv")

gene_info_file <- file.path(projectDir, "clearBoxRawData/gene_id_vs_symbol.csv")
metadata_quality_infile <- file.path(projectDir, "clearBoxCleanData/metadata_quality.csv")

cibersortx_proportion_file <- file.path(projectDir, "cibersortx/outdir/CIBERSORTx_Adjusted.txt")
cibersortx_sigMatrix_file <- file.path(projectDir, "cibersortx/outdir/CIBERSORTx_sigmatrix_Adjusted.txt")

outDesignMatrix_file <- file.path(projectDir, "clearBoxRawData", paste0(expt_name, "_design.csv"))

outdir <- file.path(projectDir, paste0(expt_name, "_output"))

dir.create(outdir)

# set max digits
options(digits=3)

# set filtering criteria
filterCPM <- 2
numSamples <- 5

```

# Import all input data tables

```{r}

## DGEList object
dge_list_obj <- readRDS(dge_list_obj_file)

## Gene ID versus GeneSymbol
gene_info <- read_csv(file = gene_info_file)

## Gene ID versus GeneSymbol
gene_info <- read_csv(file = gene_info_file)
gene_info <- gene_info[!duplicated(gene_info$ensembl_gene_id), ]

# CIBERSORTx signature matrix and proportion files
cibersortx_proportion <- read.table(file = cibersortx_proportion_file, header = TRUE, sep = "\t") %>%
  dplyr::select(., samplename = Mixture, everything()) %>%
  dplyr::select(., -'P.value', -'Correlation', -'RMSE')
cibersortx_sigMatrix <- read.table(file = cibersortx_sigMatrix_file, header = TRUE, sep = "\t")
# remove rows that contain all zero counts in the signature matrix
cibersortx_sigMatrix <- cibersortx_sigMatrix[rowSums(cibersortx_sigMatrix[,-1])>0,]

# DE tables (no covariate)
allTable_female_noCov <- read_csv(file = allTable_female_file)
allTable_male_noCov <- read_csv(file = allTable_male_file)

```

## Add cibersort proportions to metadata
```{r}
# match the order of samplename IDs
proportion_cols <- cibersortx_proportion[match(colnames(dge_list_obj), cibersortx_proportion$samplename), ]
dge_list_obj$samples <- cbind(dge_list_obj$samples, proportion_cols[, !colnames(proportion_cols) %in% "samplename", drop = FALSE])

```

## Descriptives per group

```{r}
group_by(dge_list_obj$samples, sex_outcome) %>% 
summarise(
  count = n(),
  median_birthweight = median(Birthweight, na.rm = TRUE),
  iqr_birthweight = IQR(Birthweight, na.rm = TRUE))
```

# Differential Expression
- cell-type proportions as covariate   
  - Fetal Cytotrophoblast  
  - Maternal CD8+ T-Cells  
 
```{r}
# # set the new factor levels
sex_outcome_levels <- c("F_Control", "M_Control", "F_PE", "M_PE")
#Create and write out design matrix
model_cibersortx_Covariate <- '~0 + sex_outcome + Fetal.Cytotrophoblasts'
# establish the design matrix
design_sex_PE_cibersortx <- model.matrix(formula(model_cibersortx_Covariate),
                              data = dge_list_obj$samples)
# make the column names a little nicer
colnames(design_sex_PE_cibersortx) <- c(sex_outcome_levels, "Fetal.Cytotrophoblasts")

# save a csv of the design matrix
write_csv(data.frame(design_sex_PE_cibersortx) %>%
            tibble::rownames_to_column("samplename"),
          file = outDesignMatrix_file,
          col_names = TRUE)
```

```{r}
# perform the voom
voom_sex_PE_cibersortx <- voom(dge_list_obj, design_sex_PE_cibersortx, plot = TRUE)

# fit the linear model
fit_sex_PE_cibersortx <- lmFit(voom_sex_PE_cibersortx, design_sex_PE_cibersortx)
```

```{r}
# set the contrasts - this makes it easier to "see" what we're testing
contrast_sex_PE_cibersortx <- makeContrasts(female_PE = F_PE-F_Control,
                                     male_PE = M_PE-M_Control,
                                     levels = design_sex_PE_cibersortx)
contrast_sex_PE_cibersortx

# fit a linear regression to the contrast questions
contrast_fit_sex_PE_cibersortx <- contrasts.fit(fit_sex_PE_cibersortx, contrast_sex_PE_cibersortx)

# perform bayesian adjustment
contrast_fit_sex_PE_cibersortx <- eBayes(contrast_fit_sex_PE_cibersortx)
```

```{r}
# summary table of the sex PE fit
summary(decideTests(contrast_fit_sex_PE_cibersortx,
                    adjust.method = "fdr",
                    p.value = 0.05))

# all DE results for the female_PE comparison
allTable_female_PE_cibersortx <- topTable(contrast_fit_sex_PE_cibersortx,
                             coef = 1,
                             n = Inf,
                             sort = "p")
# create output files and save
write_csv(allTable_female_PE_cibersortx,
          file = file.path(outdir, "allTable_female.csv"))


# all DE results for the male comparison
allTable_male_PE_cibersortx <- topTable(contrast_fit_sex_PE_cibersortx,
                             coef = 2,
                             n = Inf,
                             sort = "p")

write_csv(allTable_male_PE_cibersortx,
          file = file.path(outdir, "allTable_male.csv"))

dplyr::left_join(allTable_male_PE_cibersortx[, c("ensembl_gene_id", "logFC")] %>%
                   dplyr::select(ensembl_gene_id, male_logFC = logFC),
                 allTable_female_PE_cibersortx[, c("ensembl_gene_id", "logFC")] %>%
                   dplyr::select(ensembl_gene_id, female_logFC = logFC),
                 by = "ensembl_gene_id") %>%
ggplot(aes(x = male_logFC, y = female_logFC)) +
  geom_point() +
  xlim(min(allTable_male_PE_cibersortx$logFC, allTable_female_PE_cibersortx$logFC), max(allTable_male_PE_cibersortx$logFC, allTable_female_PE_cibersortx$logFC)) +
  ylim(min(allTable_male_PE_cibersortx$logFC, allTable_female_PE_cibersortx$logFC), max(allTable_male_PE_cibersortx$logFC, allTable_female_PE_cibersortx$logFC)) +
  geom_abline(intercept = 0, slope = 1, size = 0.5, colour="red")

```

## Volcano plot for visualisation of differential expression
### male

```{r volcano plots male}

decide <- decideTests(contrast_fit_sex_PE_cibersortx)
d2 <- topTable(contrast_fit_sex_PE_cibersortx, coef = 2, n = Inf, sort = "p") #[,c(2,6)]
d2$threshold <- 0
d2$threshold[0:nrow(filter(allTable_male_PE_cibersortx, adj.P.Val < 0.05))] <- 1
d2$threshold <- as.factor(d2$threshold)
d3 <- decideTests(contrast_fit_sex_PE_cibersortx) %>%
  as.data.frame()
d3 <- d3[2]
d4 <- left_join(tibble::rownames_to_column(d2), (tibble::rownames_to_column(d3)),
                       by = "rowname")
d4$male_PE <- as.factor(d4$male_PE)
d4$neg.log10FDR <- -log10(d4$adj.P.Val)
# set colours for the volcano plot
colour <- c("red", "black", "darkgreen")
FDR_FC_Decide <- decideTests(contrast_fit_sex_PE_cibersortx, lfc = 0)[,"male_PE"] %>%
  as.data.frame() %>%
  set_colnames("FDR_FC_Decide") %>%
  tibble::rownames_to_column() %>%
  left_join(., d4, by = "rowname")
FDR_FC_Decide$FDR_FC_Decide <- as.factor(FDR_FC_Decide$FDR_FC_Decide)

volcano_male <- ggplot(data = FDR_FC_Decide,
                  aes(x = logFC,
                      y = neg.log10FDR,
                      label = rowname,
                      colour = FDR_FC_Decide)) +
  geom_point(alpha=0.9, size=2.00) +
  xlab(expression(Log[2]*" Fold Change")) + ylab(expression("-ve "*Log[10]*" FDR")) +
  scale_color_manual(values = colour, name="mRNA\nRegulation",
                         breaks = c("-1", "0", "1"),
                         labels = c("Down-regulated", "Not Significant", "Up-regulated")) +
  theme(plot.title = element_text(hjust = 0.5)) +
  geom_hline(yintercept = -log10(.05), linetype = "dotted") +
  geom_vline(xintercept = c(-1, 1), linetype = "dotted") +
  theme_bw() +
  theme(text = element_text(size=18)) +
  xlim(-8, 8) +
  theme(axis.text.x = element_text(colour = "black",
                                   face = "bold"),
        axis.text.y = element_text(colour = "black",
                                   face = "bold")) +
  theme(axis.title.x = element_text(colour = "black",
                                   face = "bold"),
        axis.title.y = element_text(colour = "black",
                                   face = "bold")) +
  geom_label_repel(data = dplyr::filter(FDR_FC_Decide, FDR_FC_Decide==1),
                  aes(label = hgnc_symbol),
                  fill = "white",
                  xlim = c(8, 0),
                  max.overlaps = Inf,
                  show.legend = FALSE) +
  geom_label_repel(data = dplyr::filter(FDR_FC_Decide, FDR_FC_Decide==-1),
                  aes(label = hgnc_symbol),
                  fill = "white",
                  xlim = c(-8, 0),
                  max.overlaps = Inf,
                  show.legend = FALSE) +
  theme(legend.position="bottom")

ggsave(plot = volcano_male,
       filename = file.path(outdir, "volcano_male_cibersortx.pdf"),
       units = "in",
       width = 12,
       height = 12,
       dpi = 300)


```

### female

```{r volcano plots female}

decide <- decideTests(contrast_fit_sex_PE_cibersortx)
d2 <- topTable(contrast_fit_sex_PE_cibersortx, coef = 1, n = Inf, sort = "p") #[,c(2,6)]
d2$threshold <- 0
d2$threshold[0:nrow(filter(allTable_female_PE_cibersortx, adj.P.Val < 0.05))] <- 1
d2$threshold <- as.factor(d2$threshold)
d3 <- decideTests(contrast_fit_sex_PE_cibersortx) %>%
  as.data.frame()
d3 <- d3[1]
d4 <- left_join(tibble::rownames_to_column(d2), (tibble::rownames_to_column(d3)),
                       by = "rowname")
d4$female_PE <- as.factor(d4$female_PE)
d4$neg.log10FDR <- -log10(d4$adj.P.Val)
# set colours for the volcano plot
colour <- c("red", "black", "darkgreen")
FDR_FC_Decide <- decideTests(contrast_fit_sex_PE_cibersortx, lfc = 0)[,"female_PE"] %>%
  as.data.frame() %>%
  set_colnames("FDR_FC_Decide") %>%
  tibble::rownames_to_column() %>%
  left_join(., d4, by = "rowname")
FDR_FC_Decide$FDR_FC_Decide <- as.factor(FDR_FC_Decide$FDR_FC_Decide)

volcano_female <- ggplot(data = FDR_FC_Decide,
                  aes(x = logFC,
                      y = neg.log10FDR,
                      label = rowname,
                      colour = FDR_FC_Decide)) +
  geom_point(alpha=0.9, size=2.00) +
  xlab(expression(Log[2]*" Fold Change")) + ylab(expression("-ve "*Log[10]*" FDR")) +
  scale_color_manual(values = colour, name="mRNA\nRegulation",
                         breaks = c("-1", "0", "1"),
                         labels = c("Down-regulated", "Not Significant", "Up-regulated")) +
  theme(plot.title = element_text(hjust = 0.5)) +
  geom_hline(yintercept = -log10(.05), linetype = "dotted") +
  geom_vline(xintercept = c(-1, 1), linetype = "dotted") +
  theme_bw() +
  theme(text = element_text(size=18)) +
  theme(axis.text.x = element_text(colour = "black",
                                   face = "bold"),
        axis.text.y = element_text(colour = "black",
                                   face = "bold")) +
  theme(axis.title.x = element_text(colour = "black",
                                   face = "bold"),
        axis.title.y = element_text(colour = "black",
                                   face = "bold")) +
  geom_label_repel(data = dplyr::filter(FDR_FC_Decide, FDR_FC_Decide==1),
                  aes(label = hgnc_symbol),
                  fill = "white",
                  xlim = c(7, 0),
                  max.overlaps = Inf,
                  show.legend = FALSE) +
  geom_label_repel(data = dplyr::filter(FDR_FC_Decide, FDR_FC_Decide==-1),
                  aes(label = hgnc_symbol),
                  fill = "white",
                  xlim = c(-7, 0),
                  max.overlaps = Inf,
                  show.legend = FALSE) +
  theme(legend.position="bottom")

ggsave(plot = volcano_female,
       filename = file.path(outdir, "volcano_female_cibersortx.pdf"),
       units = "in",
       width = 12,
       height = 12,
       dpi = 300)


```

# Compare DE without covariate and with cibersortx proportions as covariate

```{r}

male_logFC <- dplyr::left_join(allTable_male_PE_cibersortx[, c("ensembl_gene_id", "logFC")] %>%
                                 dplyr::select(ensembl_gene_id, male_cibersortx_logFC = logFC),
                               allTable_male_noCov[, c("ensembl_gene_id", "logFC")] %>%
                                 dplyr::select(ensembl_gene_id, male_no_covariate_logFC = logFC),
                               by = "ensembl_gene_id") %>%
  ggplot(aes(x = male_cibersortx_logFC, y = male_no_covariate_logFC)) +
  geom_hline(yintercept=0, linetype="dashed", color = "red") +
  geom_vline(xintercept=0, linetype="dashed", color = "red") +
  geom_point() +
  xlim(-3.5, 6.6) +
  ylim(-3.5, 6.6) +
  geom_point() +
  theme_bw() +
  theme(text = element_text(size=18)) +
  theme(axis.text.x = element_text(colour = "black",
                                   face = "bold"),
        axis.text.y = element_text(colour = "black",
                                   face = "bold")) +
  theme(axis.title.x = element_text(colour = "black",
                                   face = "bold"),
        axis.title.y = element_text(colour = "black",
                                   face = "bold")) +
  ggtitle("Male DE")
ggsave(plot = male_logFC,
       filename = file.path(outdir, "male_logFC.pdf"),
       units = "in",
       width = 12,
       height = 12,
       dpi = 300)


female_logFC <- dplyr::left_join(allTable_female_PE_cibersortx[, c("ensembl_gene_id", "logFC")] %>%
                                 dplyr::select(ensembl_gene_id, female_cibersortx_logFC = logFC),
                               allTable_female_noCov[, c("ensembl_gene_id", "logFC")] %>%
                                 dplyr::select(ensembl_gene_id, female_no_covariate_logFC = logFC),
                               by = "ensembl_gene_id") %>%
  ggplot(aes(x = female_cibersortx_logFC, y = female_no_covariate_logFC)) +
  geom_hline(yintercept=0, linetype="dashed", color = "red") +
  geom_vline(xintercept=0, linetype="dashed", color = "red") +
  geom_point() +
  xlim(-3.5, 6.6) +
  ylim(-3.5, 6.6) +
  geom_point() +
  theme_bw() +
  theme(text = element_text(size=18)) +
  theme(axis.text.x = element_text(colour = "black",
                                   face = "bold"),
        axis.text.y = element_text(colour = "black",
                                   face = "bold")) +
  theme(axis.title.x = element_text(colour = "black",
                                   face = "bold"),
        axis.title.y = element_text(colour = "black",
                                   face = "bold")) +
  ggtitle("Female DE")
ggsave(plot = female_logFC,
       filename = file.path(outdir, "female_logFC.pdf"),
       units = "in",
       width = 12,
       height = 12,
       dpi = 300)

```

# Heatmaps
```{r}
# Modify ordering of the clusters using clustering callback option
callback = function(hc, mat){
    sv = svd(t(mat))$v[,1]
    dend = reorder(as.dendrogram(hc), wts = sv)
    as.hclust(dend)
}

annotation_cols <- metadata_quality[, c("samplename", "Sex", "Outcome")] %>%
  tibble::column_to_rownames("samplename")

scaled_exp <- scale(t(subset(log2CPM, rownames(log2CPM) %in% dplyr::filter(FDR_FC_Decide, threshold == 1)$ensembl_gene_id)))

# png(file=file.path(outdir, "heatmap_scaled_sig_AdjPVal_M.png"),
#     width=12,
#     height=8,
#     units = "in",
#     res = 150)
pheatmap(t(scaled_exp),
         annotation_col = annotation_cols,
         show_rownames = FALSE,
         show_colnames = TRUE,
         cutree_cols = 3,
         clustering_callback = callback)
# dev.off()

```

# Session information

```{r session info}
sessionInfo()
```
