---
title: "male_rawCounts_QC_PCA"
author: "Melanie Smith"
date: "20 June 2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    echo = TRUE,
    message = FALSE,
    warning = FALSE,
    cache = FALSE,
    fig.align = "center",
    results = "hide"
)

```

# Load required libraries

```{r}
library(plyr)
library(reshape2)
library(dplyr)
library(stringr)
library(edgeR)
library(readxl)
library(readr)
library(magrittr)
library(FactoMineR)
library(factoextra)
library(corrplot)


# set project directory
projectDir <- "/home/smit1924/sexInformedPreeclampsia"
# projectDir <- "/Users/pillmaka/Desktop/Melanie/drive-download"

expt_name <- "PCA"

log2CPM_infile <- file.path(projectDir, "/melanie_v2_0_3_output/melanie_v2_0_3logCPM.csv")
metadata_infile <- file.path(projectDir, "clearBoxRawData/cleanSampleMetadata.csv")
gene_info_file <- file.path(projectDir, "clearBoxRawData/gene_id_vs_symbol.csv")  

outdir <- file.path(projectDir, paste0(expt_name, "_output"))
# output_spreadsheet_file <- file.path(outdir, paste0(expt_name, "_.csv"))

dir.create(outdir)

```

## Data File Import
### Import log2CPM counts

```{r}
# import the counts table
log2CPM <- read_csv(file = log2CPM_infile) %>%
  as.data.frame() %>%
  tibble::column_to_rownames("ensembl_gene_id")
dim(log2CPM)
```

### Import metadata

```{r}
# import the metadata table
metadata <- read_csv(file = metadata_infile) %>%
  as.data.frame()
dim(metadata)

# ggplot(metadata_males, aes(x = Outcome, y = BMI, fill = Outcome)) + geom_boxplot() +
#   geom_point(position=position_jitterdodge())
```

### Import gene information

```{r}
## Gene ID versus GeneSymbol
gene_info <- read_csv(file = gene_info_file)
```

### subset counts and metadata

```{r}
# subset to only keep male data
metadata_males <- dplyr::filter(metadata, Sex == "M")
log2CPM_males <- dplyr::select(log2CPM, metadata_males$samplename)
```

### Drop least variable genes
This step allows me to see which genes are the most variable between all of my samples
I will subset the top 1000 most variable genes based on both I90PR (10-90) and IQR (25-75).  
First we calculate the interdecile range and plot the log2 gene expression of the top 1000 variable genes  

```{r}

#Select most variable genes
percentile_data <- apply(as.matrix(log2CPM_males), 1, quantile, c(0, .1, .25, .5, .75, .9, 1))
percentile_data[, 1:5]

# calculate the interdecile range
# These data represent the range of expression values between he lower and upper 10% of the distribution for each gene
I90PR <- percentile_data["90%", ] - percentile_data["10%", ]

gene_i90pr <- sort(I90PR, decreasing=TRUE)
# and top 5000 genes were selected.
top_genes <- names(gene_i90pr[1:1000])

top_genes_i90pr <- log2CPM_males %>%
  as.data.frame() %>% 
  tibble::rownames_to_column(var = 'gene') %>%
  dplyr::filter(gene %in% top_genes) %>%
  tibble::column_to_rownames(var = 'gene') %>%
  as.matrix()

heatmap_df_i90pr <- t(scale(t(top_genes_i90pr),
                      center = TRUE,
                      scale = TRUE))


# set up the column annotations
col_anno <- data.frame(metadata_males) %>% 
  dplyr::select(., samplename, Outcome, cohort, SmokingStatus) %>%
  tibble::column_to_rownames("samplename")


par(mar=c(6,2,2,1)+0.1) 
# png(filename=file.path(paste0(outdir, '/heatmap_i90pr_CPM.png')), width=800, height=1000)
pheatmap::pheatmap(heatmap_df_i90pr,
                   annotation_col = col_anno,
                   show_rownames = FALSE,
                   scale = "none",
                   cluster_rows = TRUE,
                   cluster_cols = TRUE,
                   main = "i90pr male")
# graphics.off()
```


```{r}

# calculate the interquartile range
# These data represent the range of expression values between he lower and upper 25% of the distribution of genes
IQR <- percentile_data["75%", ] - percentile_data["25%", ]
# sort highest to lowest expression
gene_iqr <- sort(IQR, decreasing = TRUE)
# and top 1000 genes were selected.
top_genes_iqr <- names(gene_iqr[1:1000])
 
top_genes_df_iqr <- log2CPM_males %>%
  as.data.frame() %>% 
  tibble::rownames_to_column(var = 'gene') %>%
  dplyr::filter(gene %in% top_genes_iqr) %>%
  tibble::column_to_rownames(var = 'gene') %>%
  as.matrix()

heatmap_df_iqr <- t(scale(t(top_genes_df_iqr),
                      center=TRUE,
                      scale=TRUE))

# plot the heatmap
par(mar=c(6,2,2,1)+0.1) 
# png(filename=file.path(paste0(outdir, '/heatmap_iqr_CPM.png')), width=800, height=1000)
pheatmap::pheatmap(heatmap_df_iqr,
                   annotation_col = col_anno,
                   show_rownames = FALSE,
                   scale = "none",
                   cluster_rows = TRUE,
                   cluster_cols = TRUE,
                   main = "iqr male")
# graphics.off()

```

### PCA
Lets take a look at the top 2000 genes (IQR) in a PCA.  

```{r}

# vector of top 2000 genes in the IQR
top_2000_genes_iqr <- names(gene_iqr[1:2000])

scaled_log2CPM_male <- t(scale(t(log2CPM_males),
                      center = TRUE,
                      scale = TRUE))

# PCA with top 5000 genes, all male samples
res_2000_iqr <- FactoMineR::PCA(
  scaled_log2CPM_male %>%
    as.data.frame() %>%
    tibble::rownames_to_column(var='gene') %>%
    dplyr::filter(gene %in% top_2000_genes_iqr) %>%
    tibble::column_to_rownames(var='gene') %>%
    as.matrix() %>%
    t(),
  ncp = 50,
  graph = FALSE
)

print(res_2000_iqr)

# print summary of results
summary(res_2000_iqr,
        nbelements = Inf)

# add the supplementary variables to the logged counts table
lcpm_supp_male <- scaled_log2CPM_male %>%
    as.data.frame() %>%
    tibble::rownames_to_column(var='gene') %>%
  # keep only top 2000 genes
    dplyr::filter(gene %in% top_2000_genes_iqr) %>%
    tibble::column_to_rownames(var='gene') %>%
    as.matrix() %>%
    t() %>%
  as.data.frame() %>%
  tibble::rownames_to_column("samplename") %>%
  # add the supplementary variables
  dplyr::left_join(., dplyr::select(metadata_males, samplename, GestationalAge, maternalAge, BMI, Birthweight,
                                    cohort, Outcome, Ethnicity, SmokingStatus),
                   by = "samplename") %>%
  # move samplename back into the rownames
  tibble::column_to_rownames("samplename")

# PCA with top 2000 genes, all samples and added supp variables
res_supp <- FactoMineR::PCA(lcpm_supp_male,
                            # add columns for quantitative variables
                            quanti.sup = 2001:2004,
                            # add columns for qualitative variables
                            quali.sup = 2005:2008,
                            ncp = 50,
                            graph = FALSE)
# print summary of results for all individuals
summary(res_supp, nbelements = Inf)

# description of the dimensions. Measures the correlation (R2 value) between each variable and the dimension listed. By default only correlations not equal to zero are shown. Default p-value < 0.05.
dimdesc(res_supp)

# plot PCA for indivuals
plot(res_supp,
     cex = 0.8,
     invisible = "quali",
     title = "Individuals PCA Graph")
 
plot(res_supp,
     cex = 0.8,
     habillage = "cohort")

# plot PCA with 95% confidence elipse
# png(filename=file.path(paste0(outdir, '/plotellipses_2000_iqr.png')), width=1000, height=800)
plotellipses(res_supp,
             axes = c(1, 2))
# graphics.off()

# plot PCA with 95% confidence elipse
# png(filename=file.path(paste0(outdir, '/plotellipses_2000_iqr_dim2_3.png')), width=1000, height=800)
plotellipses(res_supp,
             axes = c(2, 3))
# graphics.off()
# plot PCA with 95% confidence elipse
# png(filename=file.path(paste0(outdir, '/plotellipses_2000_iqr_dim3_4.png')), width=1000, height=800)
plotellipses(res_supp,
             axes = c(3, 4))
# graphics.off()
```

**Observation**: PE:Control seems to sit between Dim 1 and Dim 2.  
**Observation**: Cohort sit on Dim 1.  
**Observation**: Ethnicity sits on Dim 2 however, there is only one woman with Asian ethnicity so we can’t estimate the effect of ethnicity.  
**Observation**: Note that there are only two smokers - not much we can do with that.
**Observation**: Outlier samples seem to be even more extreme on the deepter PCs.  

```{r, echo = FALSE}

# fviz_pca_biplot(res_supp, 
#   habillage = "Outcome",
#   addEllipses = TRUE,
#   col.var = "red",
#   alpha.var ="cos2",
#   label = "var",
#   title = "PCA - Biplot - IQR 2000") +
#   scale_color_brewer(palette="Dark2")+
#   theme_minimal()

# plot PCA for dim1:2
plot(res_supp,
     choix = "ind",
     cex = 0.8,
     habillage = "cohort",
     title = "Individual PCA (dim 1:2)",
     axes = 1:2)

```
**Observation**: STP0596 is positioned away from the other STOP samples. This pregnancy is PE + SGA. Is this driving PC1 and PC2?  
```{r, echo=FALSE}
# plot PCA for dim2:3
plot(res_supp,
     choix = "ind",
     cex = 0.8,
     habillage = "cohort",
     title = "Individual PCA (dim 2:3)",
     axes = 2:3)
    
```

```{r, echo=FALSE}
     # plot PCA for dim3:4
plot(res_supp,
     choix = "ind",
     cex = 0.8,
     habillage = "cohort",
     title = "Individual PCA (dim 3:4)",
     axes = 3:4)
```

```{r, echo=FALSE}

# plot(res_supp,
#      choix = "var",
#      cex = 0.8,
#      title = "Variables PCA dim3:4",
#      axes = 3:4)

# which individuals contribute most to the PCs
# plot(res_supp,
#      cex = 0.8,
#      habillage = "Outcome",
#      select = "contrib 5")

# which variables contribute the most to the PCs
# plot(res_supp,
#      choix = "var",
#      select = "contrib 5")

# Extract eigenvalues/variances
get_eig(res_supp)

# Visualize eigenvalues/variances as scree plot
fviz_screeplot(res_supp,
               addlabels = TRUE,
               ylim = c(0, 50))
```

```{r, echo=FALSE}
# Extract the results for individuals
ind <- get_pca_ind(res_supp)
ind

# Coordinates of individuals
head(ind$coord)

# Graph of variables
# get the pca variables
var <- get_pca_var(res_supp)

# pull out the cos2 data
cos2_data <- data.frame(var$cos2) %>%
  dplyr::arrange(desc(Dim.1)) %>%
  # add gene name
  tibble::rownames_to_column("ensembl_gene_id") %>%
  dplyr::left_join(., gene_info, by = "ensembl_gene_id") %>%
  tidyr::drop_na(.) %>%
  tibble::column_to_rownames("hgnc_symbol") %>%
  dplyr::select(-ensembl_gene_id)

# plot cos2 (dim 1:5)
# png(filename=file.path(paste0(outdir, '/cos2_dim1_3.png')), width=800, height=150)
corrplot(t(cos2_data[1:50, 1:3]),
         is.corr=FALSE,
         title = "cos2 Dim 1:3",
         mar = c(1,1,1,1))
# graphics.off()

# plot cos2 (dim 2:5)
# png(filename=file.path(paste0(outdir, '/cos2_dim2_21.png')), width=800, height=400)
corrplot(t(cos2_data[1:50, -1]),
         is.corr=FALSE,
         title = "cos2 Dim 2:21",
         mar = c(1,1,1,1))
# graphics.off()
```

```{r, echo=FALSE}
# pull out the contribution data
contrib_data <- data.frame(var$contrib) %>%
  dplyr::arrange(desc(Dim.1)) %>%
  # add gene name
  tibble::rownames_to_column("ensembl_gene_id") %>%
  dplyr::left_join(., gene_info, by = "ensembl_gene_id") %>%
  tidyr::drop_na(.) %>%
  tibble::column_to_rownames("hgnc_symbol") %>%
  dplyr::select(-ensembl_gene_id)

# plot contribution (dim 1:5)
# png(filename=file.path(paste0(outdir, '/contrib_dim1_3.png')), width=800, height=150)
corrplot(t(contrib_data[1:50, 1:3]),
         is.corr=FALSE,
         title = "contribution Dim 1:3",
         mar = c(1,1,1,1))

# plot contribution (dim 2:5)
# png(filename=file.path(paste0(outdir, '/contrib_dim2_21.png')), width=800, height=400)
corrplot(t(contrib_data[1:50, -1]),
         is.corr=FALSE,
         title = "contribution Dim 2:21",
         mar = c(1,1,1,1))
# graphics.off()
```

```{r, echo=FALSE}

# plot cos2 qualitative variables
png(filename=file.path(paste0(outdir, '/cos2Quali_dim1_20.png')), width=400, height=800)
corrplot(t(res_supp$quali.sup$cos2),
         is.corr = FALSE,
         title = "cos2 qualitative variables",
         mar = c(1,1,1,1))
```
**Observation**: Seems to be a strong signal for PE:Control on Dim 1 but cohort is also here. How to remove the cohort signal but not the pathology??  

```{r, echo=FALSE}

# plot cos2 quantitative variables
png(filename=file.path(paste0(outdir, '/cos2Quanti_dim1_20.png')), width=400, height=800)
corrplot(t(res_supp$quanti.sup$cos),
         is.corr = FALSE,
         title = "cos2 quantitative variables",
         mar = c(1,1,1,1))

```
**Observation**: Maternal Age has a strong cos2 on Dim 1 but this is complicated because the risk associated with maternal age is a "U-shaped" one. Much younger and much older women are most at risk of developing PE.  
**Observation**: I thought I’d plot the maternal age across outcome to see if there was a pattern. Small numbers but there doesn’t appear to be a bias.  
**Observation**: Maternal BMI is often given as another risk factor for PE. We don't appear to have a bias in high or low BMI associated with outcome.  

```{r, echo=FALSE}

# get the individual contributions
# plot individual contributions to PCs
ind <- get_pca_ind(res_supp)
png(filename=file.path(paste0(outdir, '/indivCos2_dim1_20.png')), width=400, height=400)
corrplot(t(ind$cos2),
         is.corr = FALSE,
         title = "cos2 of Individuals",
         mar = c(1,1,1,1))
# graphics.off()
# png(filename=file.path(paste0(outdir, '/indivContrib_dim1_20.png')), width=400, height=400)
corrplot(t(ind$contrib),
         is.corr = FALSE,
         title = "Contribution of Individuals",
         mar = c(1,1,1,1))
# graphics.off()
```

**Observation**: SCP4010 doesn't seem to have a strong relationship with Dim 1. This is the small library.  

### Session information

```{r, sessionInfo}

sessionInfo()

```
