---
title: "PCA_RNA_quality_conc"
author: "Melanie Smith"
date: "17 January 2024"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    echo = TRUE,
    message = FALSE,
    warning = FALSE,
    cache = FALSE,
    fig.align = "center",
    results = "hide"
)

```

# Load required libraries

```{r}
library(plyr)
library(reshape2)
library(dplyr)
library(stringr)
library(edgeR)
library(readxl)
library(readr)
library(magrittr)
library(FactoMineR)
library(factoextra)
library(corrplot)


# set project directory
projectDir <- "/home/smit1924/preeclampsia_sex_chromosome_informed"
# projectDir <- "/Users/pillmaka/Desktop/Melanie/drive-download"

expt_name <- "PCA_qc"

log2CPM_infile <- file.path(projectDir, "clearBoxRawData/melanie_v2_0_3logCPM.csv")
metadata_infile <- file.path(projectDir, "clearBoxRawData/cleanSampleMetadata.csv")
gene_info_file <- file.path(projectDir, "clearBoxRawData/gene_id_vs_symbol.csv")  
rnaQuality_infile <- file.path(projectDir, "clearBoxRawData/rin_conc.xlsx")

outdir <- file.path(projectDir, paste0(expt_name, "_output"))
# output_spreadsheet_file <- file.path(outdir, paste0(expt_name, "_.csv"))

dir.create(outdir)

```

## Data File Import
### Import log2CPM counts

```{r}
# import the counts table
log2CPM <- read_csv(file = log2CPM_infile) %>%
  as.data.frame() %>%
  tibble::column_to_rownames("ensembl_gene_id")
dim(log2CPM)
```

### Import metadata and RNA quality

```{r}
# import the metadata table
metadata <- read_csv(file = metadata_infile) %>%
  as.data.frame()
dim(metadata)

# import the RNA quality file
rna_quality <- read_excel(rnaQuality_infile)
# small function to capture the samplename
create_samplename <- function(input_column) {
  # Extract the first three letters
  first_three_letters <- substr(input_column, 1, 3)
  # Extract the numbers after the first underscore
  numbers_after_first_underscore <- substr(input_column, 5, 8)
  # Concatenate the results
  samplename <- paste0(first_three_letters, numbers_after_first_underscore)
  return(samplename)
}
# add the new samplename column
rna_quality$samplename <- create_samplename(rna_quality$sample)

# import the delivery date information
deliveryDate_combined <- readr::read_csv(file.path(projectDir, "metaData_SCOPE_STOP_output/deliveryDate_combined.csv"))

# import time to freezer information
# STOP
STOP_placentaFreezerSamples <- readr::read_csv(file.path(projectDir, "clearBoxRawData/STOP_placentaFreezerSamples.csv")) %>%
  # drop all the noisy NAs down the bottom of Dylan's spreadsheet 
  dplyr::filter(complete.cases(.)) %>%
  # add a samplename that matches the rest of my data
  dplyr::mutate(., samplename = paste0("STP", Patient_ID)) %>%
  # keep one instance of each samplename
  dplyr::distinct(samplename, notCollectThreeHrs, .keep_all = TRUE)
## coerce dates from character to date format
STOP_placentaFreezerSamples$dateSampleCollected <- as.Date(STOP_placentaFreezerSamples$dateSampleCollected, format = "%d/%m/%Y")
STOP_placentaFreezerSamples$dateSampleFrozen_80C <- as.Date(STOP_placentaFreezerSamples$dateSampleFrozen_80C, format = "%d/%m/%Y")
#SCOPE from Dylan email
SCOPE_placentaFeezerSamples <- data.frame(
  SCOPE_ID = as.numeric(c(3412,3628,3660,3780,3843,3877,3929,3954,3962,3992,4059,4073,4139,4154,4157,4164,4196,4319,4378,4536,4538,4578,4706,4726,4733,4748,4809,4913)),
  DOB =	as.Date(c('27/12/2007','15/02/2008','24/02/2008','10/03/2008','21/03/2008','20/03/2008','16/04/2008','30/04/2008','13/04/2008','31/03/2008','20/04/2008','24/06/2008','4/07/2008','25/06/2008','21/06/2008','29/06/2008','24/07/2008','24/08/2008','15/09/2008','29/10/2008','10/09/2008','8/10/2008','25/11/2008','11/12/2008','4/12/2008','29/12/2008','29/11/2008','15/01/2009'),
                format = "%d/%m/%Y"),
  Date_collected_in_lab = as.Date(c('28/12/2007','19/02/2008','28/02/2008','17/03/2008','27/03/2008','27/03/2008','21/04/2008','7/05/2008','16/04/2008','2/04/2008','21/04/2008','27/06/2008','5/07/2008','27/06/2008','NA','NA','31/07/2008','27/08/2008','16/09/2008','NA','16/09/2008','13/10/2008','25/11/2008','NA','NA','30/12/2008','NA','NA'),
                                  format = "%d/%m/%Y")) %>%
  dplyr::mutate(., samplename = paste0("SCP", SCOPE_ID)) %>%
  dplyr::select(., samplename, DOB, Date_collected_in_lab)
# combine the freezer data
combined_placentaFreezerSamples <- rbind(STOP_placentaFreezerSamples %>%
                                           dplyr::select(., samplename, deliveryDate = dateSampleCollected, dateSampleFrozen_80C),
                                         SCOPE_placentaFeezerSamples %>%
                                           dplyr::select(., samplename, deliveryDate = DOB, dateSampleFrozen_80C = Date_collected_in_lab)) %>%
  # add a column to calculate the number of days taken from delivery to freezer
  dplyr::mutate(., days_to_freezer = dateSampleFrozen_80C-deliveryDate)

```

### combine metadata and RNA quality information

```{r}
metadata_quality <- dplyr::left_join(metadata, rna_quality, by = "samplename") %>%
  dplyr::left_join(., deliveryDate_combined, by = "samplename") %>%
  # add the number of days between delivery date and today
  dplyr::mutate(., days_diff = difftime(Sys.Date(), deliveryDate, units = "days")) %>%
  # add the "time to freezer" information
  dplyr::left_join(., combined_placentaFreezerSamples, by = "samplename")
```

### plot the quality and delivery date information

```{r}

# using the original delivery date (as a proxy for how old the samples are) how does sample age relate to sample RQI
metadata_quality %>%
  ggplot(aes(x = deliveryDate,
             y = RQI,
             size = 6)) +
  geom_point(aes(shape = factor(cohort),
                 colour = Conc)) +
  facet_wrap(vars(cohort), scales = "free")

# how does RNA quality (RQI) relate to the concentration of RNA
metadata_quality %>%
  ggplot(aes(x = Conc,
             y = RQI)) +
  geom_point(aes(colour = factor(cohort))) +
  geom_smooth(aes(group = cohort,
                  colour = cohort),
              method = lm)

# use the original delivery date to calculate the age of the sample and re-plot against RQN and concentration
metadata_quality %>%
  ggplot(aes(x = as.numeric(days_diff),
             y = RQI)) +
  geom_point() +
  geom_smooth(aes(group = cohort,
                  colour = cohort),
              method = lm) +
  facet_wrap(vars(cohort), scales = "free")

# does the time to freezer (days) appear to correlate with the RQI or concentration?
metadata_quality %>%
  ggplot(aes(x = as.numeric(days_to_freezer),
             y = RQI)) +
  geom_point() +
  geom_smooth(aes(group = cohort,
                  colour = cohort),
              method = lm) +
  facet_wrap(vars(cohort), scales = "free")

metadata_quality %>%
  ggplot(aes(x = as.numeric(days_to_freezer),
             y = Conc)) +
  geom_point() +
  geom_smooth(aes(group = cohort,
                  colour = cohort),
              method = lm) +
  facet_wrap(vars(cohort), scales = "free")

```

### Import gene information

```{r}
## Gene ID versus GeneSymbol
gene_info <- read_csv(file = gene_info_file)
```

### Drop least variable genes
This step allows me to see which genes are the most variable between all of my samples
I will subset the top 1000 most variable genes based on both I90PR (10-90) and IQR (25-75).  
First we calculate the interdecile range and plot the log2 gene expression of the top 1000 variable genes  

```{r}

#Select most variable genes
percentile_data <- apply(as.matrix(log2CPM), 1, quantile, c(0, .1, .25, .5, .75, .9, 1))
percentile_data[, 1:5]

# calculate the interdecile range
# These data represent the range of expression values between he lower and upper 10% of the distribution for each gene
I90PR <- percentile_data["90%", ] - percentile_data["10%", ]

gene_i90pr <- sort(I90PR, decreasing=TRUE)
# and top 5000 genes were selected.
top_genes <- names(gene_i90pr[1:1000])

top_genes_i90pr <- log2CPM %>%
  as.data.frame() %>% 
  tibble::rownames_to_column(var = 'gene') %>%
  dplyr::filter(gene %in% top_genes) %>%
  tibble::column_to_rownames(var = 'gene') %>%
  as.matrix()

heatmap_df_i90pr <- t(scale(t(top_genes_i90pr),
                      center = TRUE,
                      scale = TRUE))


# set up the column annotations
col_anno <- data.frame(metadata_quality) %>% 
  dplyr::select(., samplename, Outcome, cohort, SmokingStatus, RQI, Conc, sex_outcome) %>%
  tibble::column_to_rownames("samplename")


par(mar=c(6,2,2,1)+0.1) 
# png(filename=file.path(paste0(outdir, '/heatmap_i90pr_CPM.png')), width=800, height=1000)
pheatmap::pheatmap(heatmap_df_i90pr,
                   annotation_col = col_anno,
                   show_rownames = FALSE,
                   scale = "none",
                   cluster_rows = TRUE,
                   cluster_cols = TRUE,
                   main = "i90pr male")
# graphics.off()
```


```{r}

# calculate the interquartile range
# These data represent the range of expression values between he lower and upper 25% of the distribution of genes
IQR <- percentile_data["75%", ] - percentile_data["25%", ]
# sort highest to lowest expression
gene_iqr <- sort(IQR, decreasing = TRUE)
# and top 1000 genes were selected.
top_genes_iqr <- names(gene_iqr[1:1000])
 
top_genes_df_iqr <- log2CPM %>%
  as.data.frame() %>% 
  tibble::rownames_to_column(var = 'gene') %>%
  dplyr::filter(gene %in% top_genes_iqr) %>%
  tibble::column_to_rownames(var = 'gene') %>%
  as.matrix()

heatmap_df_iqr <- t(scale(t(top_genes_df_iqr),
                      center=TRUE,
                      scale=TRUE))

# plot the heatmap
par(mar=c(6,2,2,1)+0.1) 
# png(filename=file.path(paste0(outdir, '/heatmap_iqr_CPM.png')), width=800, height=1000)
pheatmap::pheatmap(heatmap_df_iqr,
                   annotation_col = col_anno,
                   show_rownames = FALSE,
                   scale = "none",
                   cluster_rows = TRUE,
                   cluster_cols = TRUE,
                   main = "iqr male")
# graphics.off()

```

### PCA
Lets take a look at the top 2000 genes (IQR) in a PCA.  

```{r}

# vector of top 2000 genes in the IQR
top_2000_genes_iqr <- names(gene_iqr[1:2000])

scaled_log2CPM <- t(scale(t(log2CPM),
                      center = TRUE,
                      scale = TRUE))

# PCA with top 5000 genes, all male samples
res_2000_iqr <- FactoMineR::PCA(
  scaled_log2CPM %>%
    as.data.frame() %>%
    tibble::rownames_to_column(var='gene') %>%
    dplyr::filter(gene %in% top_2000_genes_iqr) %>%
    tibble::column_to_rownames(var='gene') %>%
    as.matrix() %>%
    t(),
  ncp = 50,
  graph = FALSE
)

print(res_2000_iqr)

# print summary of results
summary(res_2000_iqr,
        nbelements = Inf)

# add the supplementary variables to the logged counts table
lcpm_supp <- scaled_log2CPM %>%
    as.data.frame() %>%
    tibble::rownames_to_column(var='gene') %>%
  # keep only top 2000 genes
    dplyr::filter(gene %in% top_2000_genes_iqr) %>%
    tibble::column_to_rownames(var='gene') %>%
    as.matrix() %>%
    t() %>%
  as.data.frame() %>%
  tibble::rownames_to_column("samplename") %>%
  # add the supplementary variables
  dplyr::left_join(., dplyr::select(metadata_quality, samplename, GestationalAge, maternalAge, BMI, Birthweight, RQI, Conc,
                                    cohort, Outcome, Ethnicity, SmokingStatus, Sex),
                   by = "samplename") %>%
  # move samplename back into the rownames
  tibble::column_to_rownames("samplename")

# PCA with top 2000 genes, all samples and added supp variables
res_supp <- FactoMineR::PCA(lcpm_supp,
                            # add columns for quantitative variables
                            quanti.sup = 2001:2006,
                            # add columns for qualitative variables
                            quali.sup = 2007:2011,
                            ncp = 50,
                            graph = FALSE)
# print summary of results for all individuals
summary(res_supp, nbelements = Inf)

# description of the dimensions. Measures the correlation (R2 value) between each variable and the dimension listed. By default only correlations not equal to zero are shown. Default p-value < 0.05.
dimdesc(res_supp)

# plot PCA for indivuals
plot(res_supp,
     cex = 0.8,
     invisible = "quali",
     title = "Individuals PCA Graph")
 
plot(res_supp,
     cex = 0.8)

# plot PCA with 95% confidence elipse
# png(filename=file.path(paste0(outdir, '/plotellipses_2000_iqr.png')), width=1000, height=800)
plotellipses(res_supp,
             axes = c(1, 2))
# graphics.off()

# plot PCA with 95% confidence elipse
# png(filename=file.path(paste0(outdir, '/plotellipses_2000_iqr_dim2_3.png')), width=1000, height=800)
plotellipses(res_supp,
             axes = c(2, 3))
# graphics.off()
# plot PCA with 95% confidence elipse
# png(filename=file.path(paste0(outdir, '/plotellipses_2000_iqr_dim3_4.png')), width=1000, height=800)
plotellipses(res_supp,
             axes = c(3, 4))
# graphics.off()
```
