---
title: "plots, lots of plots"
author: "Melanie Smith"
date: "12 July 2023"
output: html_document
---

# Load required libraries
```{r load Libraries}
library(plyr)
library(reshape2)
library(dplyr)
library(tidyr)
library(stringr)
library(edgeR)
library(readxl)
library(readr)
library(magrittr)
library(ggplot2)
library(tibble)
# library(ggbeeswarm)
# library(ggrepel)
# library(RColorBrewer)
# library(pheatmap)
# library(ggpubr)

# set project directory
projectDir <- "/home/smit1924/sexInformedPreeclampsia"
# projectDir <- "/Users/pillmaka/Desktop/Melanie/drive-download"

# define input files
jimmy_v1_6_4_inFile <- file.path(projectDir, "jimmy_output/jimmy_DE.csv")
melanie_v1_6_4_inFile <- file.path(projectDir, "melanie_v1_6_4_output/melanie_v1_6_4_DE.csv")
melanie_v2_0_3_inFile <- file.path(projectDir, "melanie_v2_0_3_output/melanie_v2_0_3_DE.csv")

jimmy_v1_6_4_filtered_inFile <- file.path(projectDir, "jimmy_output/jimmyfiltered_DE.csv")
melanie_v1_6_4_filtered_inFile <- file.path(projectDir, "melanie_v1_6_4_output/melanie_v1_6_4filtered_DE.csv")
melanie_v2_0_3_filtered_inFile <- file.path(projectDir, "melanie_v2_0_3_output/melanie_v2_0_3filtered_DE.csv")
 
```

# Import DE files

```{r}
jimmy_v1_6_4 <- read_csv(file = jimmy_v1_6_4_inFile)
melanie_v1_6_4 <- read_csv(file = melanie_v1_6_4_inFile)
melanie_v2_0_3 <- read_csv(file = melanie_v2_0_3_inFile)

jimmy_v1_6_4_filtered <- read_csv(file = jimmy_v1_6_4_filtered_inFile)
melanie_v1_6_4_filtered <- read_csv(file = melanie_v1_6_4_filtered_inFile)
melanie_v2_0_3_filtered <- read_csv(file = melanie_v2_0_3_filtered_inFile)

```

# Combine the tables (logFC only)

```{r}
# DE using subset counts
combinedLogFC <- dplyr::full_join(jimmy_v1_6_4 %>% dplyr::select(ensembl_gene_id, hgnc_symbol, logFC_F, logFC_M),
                                  melanie_v1_6_4 %>% dplyr::select(ensembl_gene_id, hgnc_symbol, logFC_F, logFC_M),
                                  by=c('hgnc_symbol', 'ensembl_gene_id'),
                                  suffix=c("_j164", "_m164")) %>%
  dplyr::full_join(., melanie_v2_0_3 %>% dplyr::select(ensembl_gene_id, hgnc_symbol, logFC_F_m203 = logFC_F, logFC_M_m203 = logFC_M),
                   by=c('hgnc_symbol', 'ensembl_gene_id'))
# DE using filtered counts (2CPM in 5 samples)
combined_filteredLogFC <- dplyr::full_join(jimmy_v1_6_4_filtered %>% dplyr::select(ensembl_gene_id, hgnc_symbol, logFC_F, logFC_M),
                                  melanie_v1_6_4_filtered %>% dplyr::select(ensembl_gene_id, hgnc_symbol, logFC_F, logFC_M),
                                  by=c('hgnc_symbol', 'ensembl_gene_id'),
                                  suffix=c("_j164", "_m164")) %>%
  dplyr::full_join(., melanie_v2_0_3_filtered %>% dplyr::select(ensembl_gene_id, hgnc_symbol, logFC_F_m203 = logFC_F, logFC_M_m203 = logFC_M),
                   by=c('hgnc_symbol', 'ensembl_gene_id'))
```

# Combine subset and filtered tables

```{r}
# combine the subset and filtered tables
logFC_subset_filtered <- dplyr::full_join(combinedLogFC, combined_filteredLogFC,
                                          by = c("hgnc_symbol", "ensembl_gene_id"),
                                          suffix = c("_subset", "_filtered")) %>%
  # replace NA with zero
  dplyr::mutate_at(c(3:14), ~replace_na(.,0))

```

# plot subset DE
## Male and female logFC compared (internally)

```{r}
# jimmy_v1_6_4 male vs female
ggplot(data = jimmy_v1_6_4,
       aes(x = logFC_M,
           y = logFC_F)) +
  geom_point()
# melanie_v1_6_4 male vs female
ggplot(data = melanie_v1_6_4,
       aes(x = logFC_M,
           y = logFC_F)) +
  geom_point()
  # melanie_v2_0_3 male vs female
ggplot(data = melanie_v2_0_3,
       aes(x = logFC_M,
           y = logFC_F)) +
  geom_point()
```

## Jimmy & Melanie v1.6.4 logFC compared

```{r}
# v1_6_4 male
ggplot(data = combinedLogFC,
       aes(x = logFC_M_j164,
           y = logFC_M_m164)) +
  geom_point() +
  ggtitle("Male log2FC Comparison (Jimmy v1.6.4 versus Melanie v1.6.4)")
# v1_6_4 female
ggplot(data = combinedLogFC,
       aes(x = logFC_F_j164,
           y = logFC_F_m164)) +
  geom_point() +
  ggtitle("Female log2FC Comparison (Jimmy v1.6.4 versus Melanie v1.6.4)")
```

## Jimmy v1.6.4 & Melanie v2.0.3 logFC compared

```{r}
# Jimmy v1_6_4 Melanie v2.0.3 male
ggplot(data = combinedLogFC,
       aes(x = logFC_M_j164,
           y = logFC_M_m203)) +
  geom_point() +
  ggtitle("Male log2FC Comparison (Jimmy v1.6.4 versus Melanie v2.0.3)")
# Jimmy v1_6_4 Melanie v2.0.3 female
ggplot(data = combinedLogFC,
       aes(x = logFC_F_j164,
           y = logFC_F_m203)) +
  geom_point() +
  ggtitle("Female log2FC Comparison (Jimmy v1.6.4 versus Melanie v2.0.3)")
```

## Melanie v1.6.4 & Melanie v2.0.3 logFC compared

```{r}
# Melanie v1_6_4 Melanie v2.0.3 male
ggplot(data = combinedLogFC,
       aes(x = logFC_M_m164,
           y = logFC_M_m203)) +
  geom_point() +
  ggtitle("Male log2FC Comparison (Melanie v1.6.4 versus Melanie v2.0.3)")
# Melanie v1_6_4 Melanie v2.0.3 female
ggplot(data = combinedLogFC,
       aes(x = logFC_F_m164,
           y = logFC_F_m203)) +
  geom_point() +
  ggtitle("Female log2FC Comparison (Melanie v1.6.4 versus Melanie v2.0.3)")
```

# Plot filtered DE 
## Jimmy & Melanie v1.6.4 logFC compared

```{r}
# v1_6_4 male
ggplot(data = combined_filteredLogFC,
       aes(x = logFC_M_j164,
           y = logFC_M_m164)) +
  geom_point() +
  ggtitle("Male (filtered) log2FC Comparison (Jimmy v1.6.4 versus Melanie v1.6.4)")
# v1_6_4 female
ggplot(data = combined_filteredLogFC,
       aes(x = logFC_F_j164,
           y = logFC_F_m164)) +
  geom_point() +
  ggtitle("Female (filtered) log2FC Comparison (Jimmy v1.6.4 versus Melanie v1.6.4)")
```

## Jimmy v1.6.4 & Melanie v2.0.3 logFC compared

```{r}
# Jimmy v1_6_4 Melanie v2.0.3 male
ggplot(data = combined_filteredLogFC,
       aes(x = logFC_M_j164,
           y = logFC_M_m203)) +
  geom_point() +
  ggtitle("Male (filtered) log2FC Comparison (Jimmy v1.6.4 versus Melanie v2.0.3)")
# Jimmy v1_6_4 Melanie v2.0.3 female
ggplot(data = combined_filteredLogFC,
       aes(x = logFC_F_j164,
           y = logFC_F_m203)) +
  geom_point() +
  ggtitle("Female (filtered) log2FC Comparison (Jimmy v1.6.4 versus Melanie v2.0.3)")
```

## Melanie v1.6.4 & Melanie v2.0.3 logFC compared

```{r}
# Melanie v1_6_4 Melanie v2.0.3 male
ggplot(data = combined_filteredLogFC,
       aes(x = logFC_M_m164,
           y = logFC_M_m203)) +
  geom_point() +
  ggtitle("Male (filtered) log2FC Comparison (Melanie v1.6.4 versus Melanie v2.0.3)")
# Melanie v1_6_4 Melanie v2.0.3 female
ggplot(data = combined_filteredLogFC,
       aes(x = logFC_F_m164,
           y = logFC_F_m203)) +
  geom_point() +
  ggtitle("Female (filtered) log2FC Comparison (Melanie v1.6.4 versus Melanie v2.0.3)")
```

# Plot subset and filtered DE 
## Jimmy v1.6.4 subset and filtered logFC compared

```{r}
# Jimmy v1_6_4 male
ggplot(data = logFC_subset_filtered,
       aes(x = logFC_M_j164_subset,
           y = logFC_M_j164_filtered)) +
  geom_point() +
  ggtitle("Male (subset v filtered) log2FC Comparison (Jimmy v1.6.4)")
# Jimmy v1_6_4 female
ggplot(data = logFC_subset_filtered,
       aes(x = logFC_F_j164_subset,
           y = logFC_F_j164_filtered)) +
  geom_point() +
  ggtitle("Female (subset v filtered) log2FC Comparison (Jimmy v1.6.4)")
```

## Melanie v1.6.4 subset and filtered logFC compared

```{r}
# Melanie v1_6_4 male
ggplot(data = logFC_subset_filtered,
       aes(x = logFC_M_m164_subset,
           y = logFC_M_m164_filtered)) +
  geom_point() +
  ggtitle("Male (subset v filtered) log2FC Comparison (Melanie v1.6.4)")
# Melanie v1_6_4 female
ggplot(data = logFC_subset_filtered,
       aes(x = logFC_F_m164_subset,
           y = logFC_F_m164_filtered)) +
  geom_point() +
  ggtitle("Female (subset v filtered) log2FC Comparison (Melanie v1.6.4)")
```

## Melanie v2.0.3 subset and filtered logFC compared

```{r}
# Melanie v2_0_3 male
ggplot(data = logFC_subset_filtered,
       aes(x = logFC_M_m203_subset,
           y = logFC_M_m203_filtered)) +
  geom_point() +
  ggtitle("Male (subset v filtered) log2FC Comparison (Melanie v2_0_3)")
# Melanie v2_0_3 female
ggplot(data = logFC_subset_filtered,
       aes(x = logFC_F_m203_subset,
           y = logFC_F_m203_filtered)) +
  geom_point() +
  ggtitle("Female (subset v filtered) log2FC Comparison (Melanie v2_0_3)")
```
