---
title: "compare regression models (deduplicated; none, bmi, maternal age)"
author: "Melanie Smith"
date: "7th July 2025"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The aim of this script is to understand how much of the DE signal is coming from BMI  


# Load required libraries
```{r load Libraries, echo=FALSE}
library(plyr)
library(reshape2)
library(dplyr)
library(stringr)
library(edgeR)
library(readxl)
library(readr)
library(magrittr)
library(tibble)
library(ggplot2)
library(ggbeeswarm)
library(ggrepel)
library(cowplot)
library(pheatmap)
```

# Set directories and files
```{r}

# set project directory
projectDir <- "/home/smit1924/preeclampsia_sex_chromosome_informed"
# projectDir <- "/Users/pillmaka/Desktop/Melanie/drive-download"

# input files
# base no-covariate model
input_allTable_male <- file.path(projectDir, "deduplicated_clearBox/output/58_sample_noCovariate/allTable_male.csv")
input_allTable_female <- file.path(projectDir, "deduplicated_clearBox/output/58_sample_noCovariate/allTable_female.csv")

# BMI as covariate model
input_allTable_male_bmi <- file.path(projectDir, "deduplicated_clearBox/output/58_sample_bmiCovariate/allTable_male.csv")
input_allTable_female_bmi <- file.path(projectDir, "deduplicated_clearBox/output/58_sample_bmiCovariate/allTable_female.csv")

# Maternal age as covariate model
input_allTable_male_maternalAge <- file.path(projectDir, "deduplicated_clearBox/output/58_sample_maternalAgeCovariate/allTable_male.csv")
input_allTable_female_maternalAge <- file.path(projectDir, "deduplicated_clearBox/output/58_sample_maternalAgeCovariate/allTable_female.csv")

# BME and maternal age covariate model


metadata_file <- file.path(projectDir, "clearBoxRawData/cleanSampleMetadata.csv")

# set the output directory
outdir <- file.path(projectDir, "deduplicated_clearBox/output/compare_regression_models")

# output files

# set figure files
## No covariate compared to bmi as covariate
output_scatterplot_logFC_noCovariate_bmi_male_file <- file.path(outdir, "scatterplot_logFC_noCovariate_bmi_male.png")
output_scatterplot_adj.P.Val_noCovariate_bmi_male_file <- file.path(outdir, "scatterplot_adj.P.Val_noCovariate_bmi_male.png")
output_scatterplot_logFC_noCovariate_bmi_female_file <- file.path(outdir, "scatterplot_logFC_noCovariate_bmi_female.png")
output_scatterplot_P.Val_noCovariate_bmi_female_file <- file.path(outdir, "scatterplot_adj.P.Val_noCovariate_bmi_female.png")

## No covariate compared to maternal age as covariate
output_scatterplot_logFC_noCovariate_maternalAge_male_file <- file.path(outdir, "scatterplot_logFC_noCovariate_maternalAge_male.png")
output_scatterplot_adj.P.Val_noCovariate_maternalAge_male_file <- file.path(outdir, "scatterplot_adj.P.Val_noCovariate_maternalAge_male.png")
output_scatterplot_logFC_noCovariate_maternalAge_female_file <- file.path(outdir, "scatterplot_logFC_noCovariate_maternalAge_female.png")
output_scatterplot_P.Val_noCovariate_maternalAge_female_file <- file.path(outdir, "scatterplot_adj.P.Val_noCovariate_maternalAge_female.png")

## Save the panel plots
output_panel_plot_male_file <- file.path(outdir, "panelPlot_male.png")
output_panel_plot_female_file <- file.path(outdir, "panelPlot_female.png")

dir.create(outdir)

```

# Functions

```{r}
# updated function to pull the legend
## this is an issue with the female plots because most of the colour legend components are blank
## I found this solution on stack exchange here <https://stackoverflow.com/questions/78163631/r-get-legend-from-cowplot-package-no-longer-work-for-ggplot2-version-3-5-0>
get_legend2 <- function(plot, legend = NULL) {
  if (is.ggplot(plot)) {
    gt <- ggplotGrob(plot)
  } else {
    if (is.grob(plot)) {
      gt <- plot
    } else {
      stop("Plot object is neither a ggplot nor a grob.")
    }
  }
  pattern <- "guide-box"
  if (!is.null(legend)) {
    pattern <- paste0(pattern, "-", legend)
  }
  indices <- grep(pattern, gt$layout$name)
  not_empty <- !vapply(
    gt$grobs[indices], 
    inherits, what = "zeroGrob", 
    FUN.VALUE = logical(1)
  )
  indices <- indices[not_empty]
  if (length(indices) > 0) {
    return(gt$grobs[[indices[1]]])
  }
  return(NULL)
}
```


# Import allTable files
```{r}
# No covariate model
allTable_male <- readr::read_csv(file = input_allTable_male)
allTable_female <- readr::read_csv(file = input_allTable_female)

# BMI as covariate model
allTable_male_bmi <- readr::read_csv(file = input_allTable_male_bmi)
allTable_female_bmi <- readr::read_csv(file = input_allTable_female_bmi)

# Maternal age as covariate model
allTable_male_maternalAge <- readr::read_csv(file = input_allTable_male_maternalAge)
allTable_female_maternalAge <- readr::read_csv(file = input_allTable_female_maternalAge)
```

# Combine allTable objects male
```{r}
# Combine the allTables using descriptive column headers
combined_allTable_male <- dplyr::left_join(dplyr::select(allTable_male, ensembl, hgnc_symbol, logFC_male = logFC, adj.P.Val_male = adj.P.Val),
                                           dplyr::select(allTable_male_bmi, ensembl, hgnc_symbol, logFC_male_bmi = logFC, adj.P.Val_male_bmi = adj.P.Val),
                                           by = join_by(ensembl, hgnc_symbol)) %>%
  dplyr::left_join(., dplyr::select(allTable_male_maternalAge, ensembl, hgnc_symbol, logFC_male_maternalAge = logFC, adj.P.Val_male_maternalAge = adj.P.Val),
                   by = join_by(ensembl, hgnc_symbol))

# Create significance categories between no covariate and BMI using consensus approach
combined_allTable_male <- combined_allTable_male %>%
  mutate(
    # Define significance threshold
    sig_threshold = 0.05,
    
    # Create significance categories
    sig_category_no_bmi = case_when(
      adj.P.Val_male < sig_threshold & adj.P.Val_male_bmi < sig_threshold ~ "Both significant",
      adj.P.Val_male < sig_threshold & adj.P.Val_male_bmi >= sig_threshold ~ "Only base model significant",
      adj.P.Val_male >= sig_threshold & adj.P.Val_male_bmi < sig_threshold ~ "Only BMI covariate model significant",
      adj.P.Val_male >= sig_threshold & adj.P.Val_male_bmi >= sig_threshold ~ "Neither significant"
    ),
    
    # Factor for consistent ordering in legend
    sig_category_no_bmi = factor(sig_category_no_bmi, 
                         levels = c("Both significant", "Only base model significant", 
                                   "Only BMI covariate model significant", "Neither significant"))
  ) %>%
# Create significance categories between no covariate and maternal age using consensus approach  
    mutate(
    # Define significance threshold
    sig_threshold = 0.05,
    
    # Create significance categories
    sig_category_no_maternalAge = case_when(
      adj.P.Val_male < sig_threshold & adj.P.Val_male_maternalAge < sig_threshold ~ "Both significant",
      adj.P.Val_male < sig_threshold & adj.P.Val_male_maternalAge >= sig_threshold ~ "Only base model significant",
      adj.P.Val_male >= sig_threshold & adj.P.Val_male_maternalAge < sig_threshold ~ "Only maternal age covariate model significant",
      adj.P.Val_male >= sig_threshold & adj.P.Val_male_maternalAge >= sig_threshold ~ "Neither significant"
    ),
    
    # Factor for consistent ordering in legend
    sig_category_no_maternalAge = factor(sig_category_no_maternalAge, 
                         levels = c("Both significant", "Only base model significant", 
                                   "Only maternal age covariate model significant", "Neither significant"))
  )
  


```

## Scatter plot to compare logFC of the no covariate and BMI as covariate models male

```{r}

# Calculate axis limits to ensure square plot
axis_limits <- range(c(combined_allTable_male$logFC_male, combined_allTable_male$logFC_male_bmi), na.rm = TRUE)
axis_limits <- c(floor(axis_limits[1] * 1.1), ceiling(axis_limits[2] * 1.1))

# Create the logFC plot
logFC_noCovariate_bmi_male <- ggplot(data = combined_allTable_male, aes(x = logFC_male, y = logFC_male_bmi)) +
  
  # Add x=y reference line
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "gray50", linewidth = 0.8) +
  
  # Add all points in grey as background layer
  geom_point(alpha = 0.4, size = 1.2, color = "lightgray") +
  
  # Overlay significant points on top
  geom_point(data = filter(combined_allTable_male, sig_category_no_bmi != "Neither significant"),
             aes(color = sig_category_no_bmi), alpha = 0.8, size = 1.5) +
  
  # Custom color palette
  scale_color_manual(
    name = "Significance",
    values = c(
      "Both significant" = "#8B0000",          # Dark red - strongest evidence
      "Only base model significant" = "#FF6B6B", # Light red - Model 1 only
      "Only BMI covariate model significant" = "#4ECDC4", # Teal - Model 2 only  
      "Neither significant" = "#D3D3D3"       # Light gray - no evidence
    )
  ) +
  
  # Set equal axis limits for square plot
  xlim(axis_limits) +
  ylim(axis_limits) +
  
  # Ensure 1:1 aspect ratio
  coord_fixed(ratio = 1) +
  
  # Clean theme
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),
    legend.position = "bottom",
    legend.title = element_text(size = 12, face = "bold"),
    axis.title = element_text(size = 12, face = "bold"),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 11, hjust = 0.5, color = "gray30")
  ) +
  
  # Labels
  labs(
    x = "logFC (Male Model)",
    y = "logFC (Male + BMI Model)"
  ) +
  
  # Guide for better legend
  guides(color = guide_legend(override.aes = list(size = 3, alpha = 1)))
# save the plot output
ggsave(filename = output_scatterplot_logFC_noCovariate_bmi_male_file,
       plot = logFC_noCovariate_bmi_male,
       width = 8,
       height = 8,
       units = "in",
       dpi = 150)

```

## Scatter plot to compare adj.P.Val of the no covariate and BMI as covariate models male

```{r}

# Calculate axis limits to ensure square plot
axis_limits <- range(c(-log10(combined_allTable_male$adj.P.Val_male), -log10(combined_allTable_male$adj.P.Val_male_bmi)), na.rm = TRUE)
axis_limits <- c(floor(axis_limits[1] * 1.1), ceiling(axis_limits[2] * 1.1))

# Create the plot
adj.P.Val_noCovariate_bmi_male <- ggplot(data = combined_allTable_male, aes(x = -log10(adj.P.Val_male), y = -log10(adj.P.Val_male_bmi))) +
  
  # Add x=y reference line
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "gray50", linewidth = 0.8) +
  
  # Add all points in grey as background layer
  geom_point(alpha = 0.4, size = 1.2, color = "lightgray") +
  
  # Overlay significant points on top
  geom_point(data = filter(combined_allTable_male, sig_category_no_bmi != "Neither significant"),
             aes(color = sig_category_no_bmi), alpha = 0.8, size = 1.5) +
  
  # Custom color palette
  scale_color_manual(
    name = "Significance",
    values = c(
      "Both significant" = "#8B0000",         # Dark red - strongest evidence
      "Only base model significant" = "#FF6B6B", # Light red - Model 1 only
      "Only BMI covariate model significant" = "#4ECDC4", # Teal - Model 2 only  
      "Neither significant" = "#D3D3D3"       # Light gray - no evidence
    )
  ) +
  
  # Set equal axis limits for square plot
  xlim(axis_limits) +
  ylim(axis_limits) +
  
  # Ensure 1:1 aspect ratio
  coord_fixed(ratio = 1) +
  
  # Clean theme
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),
    legend.position = "bottom",
    legend.title = element_text(size = 12, face = "bold"),
    axis.title = element_text(size = 12, face = "bold"),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 11, hjust = 0.5, color = "gray30")
  ) +
  
  # Labels
  labs(
    x = "-log10 adjusted p-value (outcome only model)",
    y = "-log10 adjusted P-value (outcome + BMI model)"
  ) +
  
  # Guide for better legend
  guides(color = guide_legend(override.aes = list(size = 3, alpha = 1)))

# save the plot output
ggsave(filename = output_scatterplot_adj.P.Val_noCovariate_bmi_male_file,
       plot = adj.P.Val_noCovariate_bmi_male,
       width = 8,
       height = 8,
       units = "in",
       dpi = 150)

```

## Scatter plot to compare logFC of the no covariate and maternal age as covariate models male

```{r}

# Calculate axis limits to ensure square plot
axis_limits <- range(c(combined_allTable_male$logFC_male, combined_allTable_male$logFC_male_maternalAge), na.rm = TRUE)
axis_limits <- c(floor(axis_limits[1] * 1.1), ceiling(axis_limits[2] * 1.1))

# Create the logFC plot
logFC_noCovariate_maternalAge_male <- ggplot(data = combined_allTable_male, aes(x = logFC_male, y = logFC_male_maternalAge)) +
  
  # Add x=y reference line
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "gray50", linewidth = 0.8) +
  
  # Add all points in grey as background layer
  geom_point(alpha = 0.4, size = 1.2, color = "lightgray") +
  
  # Overlay significant points on top
  geom_point(data = filter(combined_allTable_male, sig_category_no_maternalAge != "Neither significant"),
             aes(color = sig_category_no_maternalAge), alpha = 0.8, size = 1.5) +
  
  # Custom color palette
  scale_color_manual(
    name = "Significance",
    values = c(
      "Both significant" = "#8B0000",         # Dark red - strongest evidence
      "Only base model significant" = "#FF6B6B", # Light red - Model 1 only
      "Only maternal age covariate model significant" = "#4ECDC4", # Teal - Model 2 only  
      "Neither significant" = "#D3D3D3"       # Light gray - no evidence
    )
  ) +
  
  # Set equal axis limits for square plot
  xlim(axis_limits) +
  ylim(axis_limits) +
  
  # Ensure 1:1 aspect ratio
  coord_fixed(ratio = 1) +
  
  # Clean theme
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),
    legend.position = "bottom",
    legend.title = element_text(size = 12, face = "bold"),
    axis.title = element_text(size = 12, face = "bold"),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 11, hjust = 0.5, color = "gray30")
  ) +
  
  # Labels
  labs(
    x = "logFC (outcome only model)",
    y = "logFC (outcome + maternal age model)"
  ) +
  
  # Guide for better legend
  guides(color = guide_legend(override.aes = list(size = 3, alpha = 1)))
# save the plot output
ggsave(filename = output_scatterplot_logFC_noCovariate_maternalAge_male_file,
       plot = logFC_noCovariate_maternalAge_male,
       width = 8,
       height = 8,
       units = "in",
       dpi = 150)

# Optional: Print summary of significance categories
cat("Summary of significance categories:\n")
print(table(combined_allTable_male$sig_category_no_maternalAge, useNA = "ifany"))
```

## Scatter plot to compare adj.P.Val of the no covariate and maternal age as covariate models male

```{r}

# Calculate axis limits to ensure square plot
axis_limits <- range(c(-log10(combined_allTable_male$adj.P.Val_male), -log10(combined_allTable_male$adj.P.Val_male_maternalAge)), na.rm = TRUE)
axis_limits <- c(floor(axis_limits[1] * 1.1), ceiling(axis_limits[2] * 1.1))

# Create the plot
adj.P.Val_noCovariate_maternalAge_male <- ggplot(data = combined_allTable_male, aes(x = -log10(adj.P.Val_male), y = -log10(adj.P.Val_male_maternalAge))) +
  
  # Add x=y reference line
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "gray50", linewidth = 0.8) +
  
  # Add all points in grey as background layer
  geom_point(alpha = 0.4, size = 1.2, color = "lightgray") +
  
  # Overlay significant points on top
  geom_point(data = filter(combined_allTable_male, sig_category_no_maternalAge != "Neither significant"),
             aes(color = sig_category_no_maternalAge), alpha = 0.8, size = 1.5) +
  
  # Custom color palette
  scale_color_manual(
    name = "Significance",
    values = c(
      "Both significant" = "#8B0000",         # Dark red - strongest evidence
      "Only base model significant" = "#FF6B6B", # Light red - Model 1 only
      "Only maternal age covariate model significant" = "#4ECDC4", # Teal - Model 2 only  
      "Neither significant" = "#D3D3D3"       # Light gray - no evidence
    )
  ) +
  
  # Set equal axis limits for square plot
  xlim(axis_limits) +
  ylim(axis_limits) +
  
  # Ensure 1:1 aspect ratio
  coord_fixed(ratio = 1) +
  
  # Clean theme
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),
    legend.position = "bottom",
    legend.title = element_text(size = 12, face = "bold"),
    axis.title = element_text(size = 12, face = "bold"),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 11, hjust = 0.5, color = "gray30")
  ) +
  
  # Labels
  labs(
    x = "-log10 adjusted p-value (outcome only model)",
    y = "-log10 adjusted P-value (outcome + maternal age model)"
  ) +
  
  # Guide for better legend
  guides(color = guide_legend(override.aes = list(size = 3, alpha = 1)))

# save the plot output
ggsave(filename = output_scatterplot_adj.P.Val_noCovariate_maternalAge_male_file,
       plot = adj.P.Val_noCovariate_maternalAge_male,
       width = 8,
       height = 8,
       units = "in",
       dpi = 150)

```

## Panel plot male

```{r}
# Step 1: Extract the legend from one of the plots (must include a legend!)
# get the legend
plot_legend_male <- get_legend2(logFC_noCovariate_bmi_male)

# Step 2: Remove legends from all individual plots
logFC_noCovariate_bmi_male_no_legend <- logFC_noCovariate_bmi_male + theme(legend.position = "none")
adj.P.Val_noCovariate_bmi_male_no_legend <- adj.P.Val_noCovariate_bmi_male + theme(legend.position = "none")
logFC_noCovariate_maternalAge_male_no_legend <- logFC_noCovariate_maternalAge_male + theme(legend.position = "none")
adj.P.Val_noCovariate_maternalAge_male_no_legend <- adj.P.Val_noCovariate_maternalAge_male + theme(legend.position = "none")

# Step 3: Create the 2x2 panel without legends
panel_only_male <- plot_grid(
  logFC_noCovariate_bmi_male_no_legend,
  adj.P.Val_noCovariate_bmi_male_no_legend,
  logFC_noCovariate_maternalAge_male_no_legend,
  adj.P.Val_noCovariate_maternalAge_male_no_legend,
  labels = c("a", "b", "c", "d"),
  label_size = 14,
  label_fontface = "bold",
  ncol = 2,
  align = "hv"
)

# Step 4: Combine panel and legend
panel_plot_male <- plot_grid(
  panel_only_male,
  plot_legend_male,
  ncol = 1,
  rel_heights = c(1, 0.1)  # Adjust as needed to balance panel and legend height
)

# Step 5: (Optional) Save the combined figure
ggsave(filename = output_panel_plot_male_file,
       plot = panel_plot_male,
       width = 11,
       height = 11,
       dpi = 300)
```


# Combine allTable objects female
```{r}
# Combine the allTables using descriptive column headers
combined_allTable_female <- dplyr::left_join(dplyr::select(allTable_female, ensembl, hgnc_symbol, logFC_female = logFC, adj.P.Val_female = adj.P.Val, P.Value_female = P.Value),
                                           dplyr::select(allTable_female_bmi, ensembl, hgnc_symbol, logFC_female_bmi = logFC, adj.P.Val_female_bmi = adj.P.Val, P.Value_female_bmi = P.Value),
                                           by = join_by(ensembl, hgnc_symbol)) %>%
  dplyr::left_join(., dplyr::select(allTable_female_maternalAge, ensembl, hgnc_symbol, logFC_female_maternalAge = logFC, adj.P.Val_female_maternalAge = adj.P.Val, P.Value_female_maternalAge = P.Value),
                   by = join_by(ensembl, hgnc_symbol))

# Create significance categories between no covariate and BMI using consensus approach
combined_allTable_female <- combined_allTable_female %>%
  mutate(
    # Define significance threshold
    sig_threshold = 0.05,
    
    # Create significance categories
    sig_category_no_bmi = case_when(
      adj.P.Val_female < sig_threshold & adj.P.Val_female_bmi < sig_threshold ~ "Both significant",
      adj.P.Val_female < sig_threshold & adj.P.Val_female_bmi >= sig_threshold ~ "Only base model significant",
      adj.P.Val_female >= sig_threshold & adj.P.Val_female_bmi < sig_threshold ~ "Only BMI covariate model significant",
      adj.P.Val_female >= sig_threshold & adj.P.Val_female_bmi >= sig_threshold ~ "Neither significant"
    ),
    
    # Factor for consistent ordering in legend
    sig_category_no_bmi = factor(sig_category_no_bmi, 
                                 levels = c("Both significant", "Only base model significant", 
                                            "Only BMI covariate model significant", "Neither significant"))
  ) %>%
  # Create significance categories between no covariate and maternal age using consensus approach  
  mutate(
    # Define significance threshold
    sig_threshold = 0.05,
    
    # Create significance categories
    sig_category_no_maternalAge = case_when(
      adj.P.Val_female < sig_threshold & adj.P.Val_female_maternalAge < sig_threshold ~ "Both significant",
      adj.P.Val_female < sig_threshold & adj.P.Val_female_maternalAge >= sig_threshold ~ "Only base model significant",
      adj.P.Val_female >= sig_threshold & adj.P.Val_female_maternalAge < sig_threshold ~ "Only maternal age covariate model significant",
      adj.P.Val_female >= sig_threshold & adj.P.Val_female_maternalAge >= sig_threshold ~ "Neither significant"
    )) %>%
  mutate(sig_category_no_bmi = factor(sig_category_no_bmi, levels = c(
    "Both significant",
    "Only base model significant",
    "Only BMI covariate model significant",
    "Neither significant"
  )))

```

## Scatter plot to compare logFC of the no covariate and BMI as a covariate models female

```{r}
# Calculate axis limits to ensure square plot
axis_limits <- range(c(combined_allTable_female$logFC_female, combined_allTable_female$logFC_female_bmi), na.rm = TRUE)
axis_limits <- c(floor(axis_limits[1] * 1.1), ceiling(axis_limits[2] * 1.1))

# Explicitly set levels in the factor to ensure all categories are present
combined_allTable_female <- combined_allTable_female %>%
  mutate(sig_category_no_bmi = factor(sig_category_no_bmi, levels = c(
    "Both significant",
    "Only base model significant",
    "Only BMI covariate model significant",
    "Neither significant"
  )))

# Define custom color palette (without newline characters)
logFC_noCovariate_bmi_female <- ggplot(data = combined_allTable_female, aes(x = logFC_female, y = logFC_female_bmi)) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "gray50", linewidth = 0.8) +
  geom_point(alpha = 0.4, size = 1.2, color = "lightgray", show.legend = FALSE) +
  geom_point(data = filter(combined_allTable_female, sig_category_no_bmi != "Neither significant"),
             aes(color = sig_category_no_bmi), alpha = 0.8, size = 1.5) +
  scale_color_manual(
    name = "Significance",
    values = c(
      "Both significant" = "#8B0000",            # Dark red
      "Only base model significant" = "#FF6B6B", # Light red
      "Only BMI covariate model significant" = "#4ECDC4", # Teal
      "Neither significant" = "#D3D3D3"          # Light gray
    ),
    limits = c("Both significant", 
               "Only base model significant", 
               "Only BMI covariate model significant", 
               "Neither significant")  # Force all categories to show
  ) +
  xlim(axis_limits) +
  ylim(axis_limits) +
  coord_fixed(ratio = 1) +
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),
    legend.position = "bottom",
    legend.title = element_text(size = 12, face = "bold"),
    axis.title = element_text(size = 12, face = "bold"),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 11, hjust = 0.5, color = "gray30")
  ) +
  labs(
    x = "logFC (outcome only model)",
    y = "logFC (outcome + BMI model)"
  ) +
  guides(color = guide_legend(override.aes = list(size = 3, alpha = 1)))

ggsave(filename = output_scatterplot_logFC_noCovariate_bmi_female_file,
       plot = logFC_noCovariate_bmi_female,
       width = 8,
       height = 8,
       units = "in",
       dpi = 150)
```

## Scatter plot to compare P-value of the no covariate and BMI as covariate models female

```{r}
# Calculate axis limits to ensure square plot
axis_limits <- range(c(-log10(combined_allTable_female$P.Value_female), -log10(combined_allTable_female$P.Value_female_bmi)), na.rm = TRUE)
axis_limits <- c(floor(axis_limits[1] * 1.1), ceiling(axis_limits[2] * 1.1))

# Create the plot
P.Val_noCovariate_bmi_female <- ggplot(data = combined_allTable_female, aes(x = -log10(P.Value_female), y = -log10(P.Value_female_bmi))) +
  
  # Add x=y reference line
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "gray50", linewidth = 0.8) +
  
  # Add all points in grey as background layer
  geom_point(alpha = 0.4, size = 1.2, color = "lightgray") +
  
  # Overlay significant points on top
  geom_point(data = filter(combined_allTable_female, sig_category_no_bmi != "Neither significant"),
             aes(color = sig_category_no_bmi), alpha = 0.8, size = 1.5) +
  
  # Custom color palette
  scale_color_manual(
    name = "Significance",
    values = c(
      "Both significant" = "#8B0000",         # Dark red - strongest evidence
      "Only base model significant" = "#FF6B6B", # Light red - Model 1 only
      "Only BMI covariate model significant" = "#4ECDC4", # Teal - Model 2 only  
      "Neither significant" = "#D3D3D3"       # Light gray - no evidence
    ),
    limits = c("Both significant", 
               "Only base model significant", 
               "Only BMI covariate model significant", 
               "Neither significant")  # Force all categories to show
  ) +
  
  # Set equal axis limits for square plot
  xlim(axis_limits) +
  ylim(axis_limits) +
  
  # Ensure 1:1 aspect ratio
  coord_fixed(ratio = 1) +
  
  # Clean theme
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),
    legend.position = "bottom",
    legend.title = element_text(size = 12, face = "bold"),
    axis.title = element_text(size = 12, face = "bold"),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 11, hjust = 0.5, color = "gray30")
  ) +
  
  # Labels
  labs(
    x = "-log10 P-value (outcome only model)",
    y = "-log10 P-value (outcome + BMI model)"
  ) +
  
  # Guide for better legend
  guides(color = guide_legend(override.aes = list(size = 3, alpha = 1)))

# save the plot output
ggsave(filename = output_scatterplot_P.Val_noCovariate_bmi_female_file,
       plot = P.Val_noCovariate_bmi_female,
       width = 8,
       height = 8,
       units = "in",
       dpi = 150)
```

## Scatter plot to compare logFC of the no covariate and maternal age as covariate models female

```{r}
# Calculate axis limits to ensure square plot
axis_limits <- range(c(combined_allTable_female$logFC_female, combined_allTable_female$logFC_female_maternalAge), na.rm = TRUE)
axis_limits <- c(floor(axis_limits[1] * 1.1), ceiling(axis_limits[2] * 1.1))

# Create the logFC plot
logFC_noCovariate_maternalAge_female <- ggplot(data = combined_allTable_female, aes(x = logFC_female, y = logFC_female_maternalAge)) +
  
  # Add x=y reference line
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "gray50", linewidth = 0.8) +
  
  # Add all points in grey as background layer
  geom_point(alpha = 0.4, size = 1.2, color = "lightgray") +
  
  # Overlay significant points on top
  geom_point(data = filter(combined_allTable_female, sig_category_no_maternalAge != "Neither significant"),
             aes(color = sig_category_no_maternalAge), alpha = 0.8, size = 1.5) +
  
  # Custom color palette
  scale_color_manual(
    name = "Significance",
    values = c(
      "Both significant" = "#8B0000",         # Dark red - strongest evidence
      "Only base model significant" = "#FF6B6B", # Light red - Model 1 only
      "Only maternal age covariate model significant" = "#4ECDC4", # Teal - Model 2 only  
      "Neither significant" = "#D3D3D3"       # Light gray - no evidence
    ),
    limits = c("Both significant", 
               "Only base model significant", 
               "Only BMI covariate model significant", 
               "Neither significant")  # Force all categories to show
  ) +
  
  # Set equal axis limits for square plot
  xlim(axis_limits) +
  ylim(axis_limits) +
  
  # Ensure 1:1 aspect ratio
  coord_fixed(ratio = 1) +
  
  # Clean theme
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),
    legend.position = "bottom",
    legend.title = element_text(size = 12, face = "bold"),
    axis.title = element_text(size = 12, face = "bold"),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 11, hjust = 0.5, color = "gray30")
  ) +
  
  # Labels
  labs(
    x = "logFC (outcome only model)",
    y = "logFC (outcome + maternal age model)"
  ) +
  
  # Guide for better legend
  guides(color = guide_legend(override.aes = list(size = 3, alpha = 1)))
# save the plot output
ggsave(filename = output_scatterplot_logFC_noCovariate_maternalAge_female_file,
       plot = logFC_noCovariate_maternalAge_female,
       width = 8,
       height = 8,
       units = "in",
       dpi = 150)
```

## Scatter plot to compare P-value of the no covariate and maternal age as covariate model female

```{r}
# Calculate axis limits to ensure square plot
axis_limits <- range(c(-log10(combined_allTable_female$P.Value_female), -log10(combined_allTable_female$P.Value_female_maternalAge)), na.rm = TRUE)
axis_limits <- c(floor(axis_limits[1] * 1.1), ceiling(axis_limits[2] * 1.1))

# Create the plot
P.Value_noCovariate_maternalAge_female <- ggplot(data = combined_allTable_female, aes(x = -log10(P.Value_female), y = -log10(P.Value_female_maternalAge))) +
  
  # Add x=y reference line
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "gray50", linewidth = 0.8) +
  
  # Add all points in grey as background layer
  geom_point(alpha = 0.4, size = 1.2, color = "lightgray") +
  
  # Overlay significant points on top
  geom_point(data = filter(combined_allTable_female, sig_category_no_maternalAge != "Neither significant"),
             aes(color = sig_category_no_maternalAge), alpha = 0.8, size = 1.5) +
  
  # Custom color palette
  scale_color_manual(
    name = "Significance",
    values = c(
      "Both significant" = "#8B0000",         # Dark red - strongest evidence
      "Only base model significant" = "#FF6B6B", # Light red - Model 1 only
      "Only maternal age covariate model significant" = "#4ECDC4", # Teal - Model 2 only  
      "Neither significant" = "#D3D3D3"       # Light gray - no evidence
    ),
    limits = c("Both significant", 
               "Only base model significant", 
               "Only BMI covariate model significant", 
               "Neither significant")  # Force all categories to show
  ) +
  
  # Set equal axis limits for square plot
  xlim(axis_limits) +
  ylim(axis_limits) +
  
  # Ensure 1:1 aspect ratio
  coord_fixed(ratio = 1) +
  
  # Clean theme
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),
    legend.position = "bottom",
    legend.title = element_text(size = 12, face = "bold"),
    axis.title = element_text(size = 12, face = "bold"),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 11, hjust = 0.5, color = "gray30")
  ) +
  
  # Labels
  labs(
    x = "-log10 adjusted p-value (outcome only model)",
    y = "-log10 adjusted P-value (outcome + maternal age model)"
  ) +
  
  # Guide for better legend
  guides(color = guide_legend(override.aes = list(size = 3, alpha = 1)))

# save the plot output
ggsave(filename = output_scatterplot_P.Val_noCovariate_maternalAge_female_file,
       plot = P.Value_noCovariate_maternalAge_female,
       width = 8,
       height = 8,
       units = "in",
       dpi = 150)
```

## Panel plot female

```{r}
# get the legend
plot_legend <- get_legend2(logFC_noCovariate_bmi_female)

# Step 2: Remove legends from all individual plots
logFC_noCovariate_bmi_female_no_legend <- logFC_noCovariate_bmi_female + theme(legend.position = "none")
P.Val_noCovariate_bmi_female_no_legend <- P.Val_noCovariate_bmi_female + theme(legend.position = "none")
logFC_noCovariate_maternalAge_female_no_legend <- logFC_noCovariate_maternalAge_female + theme(legend.position = "none")
P.Value_noCovariate_maternalAge_female_no_legend <- P.Value_noCovariate_maternalAge_female + theme(legend.position = "none")

# Step 3: Create the 2x2 panel without legends
panel_only <- plot_grid(
  logFC_noCovariate_bmi_female_no_legend,
  P.Val_noCovariate_bmi_female_no_legend,
  logFC_noCovariate_maternalAge_female_no_legend,
  P.Value_noCovariate_maternalAge_female_no_legend,
  labels = c("a", "b", "c", "d"),
  label_size = 14,
  label_fontface = "bold",
  ncol = 2,
  align = "hv"
)

# Step 4: Combine panel and legend
panel_plot_female <- plot_grid(
  panel_only,
  plot_legend,
  ncol = 1,
  rel_heights = c(1, 0.1)  # Adjust as needed to balance panel and legend height
)

# Step 5: (Optional) Save the combined figure
ggsave(filename = output_panel_plot_female_file,
       plot = panel_plot_female,
       width = 11,
       height = 11,
       dpi = 300)

```






